---
title: "Bib_study_workflow"
author: "Josefina Weinerova"
date: "2025-08-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# package loading
```{r}
# install.packages("openalexR")
if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  openalexR, cowplot, tidyverse, httr, jsonlite,
  citecorp, rcrossref, purrr, dplyr, tibble, fect, furrr, future, data.table,
  panelView
)
```

# load dataset
```{r}
df <- read_csv("data/2025-02-07_COSdata_validated_delivered.csv")
head(df)

# remove spaces from dois
df$doi_o <- gsub(" ", "", df$doi_o)
df$doi_r <- gsub(" ", "", df$doi_r)
```
# what to do if multiple replications per study
- in the dataset for analysis this will be slightly different as rows are study - replication pairs, while here rows are effects
- in analysis will select those original replications pairs so that its the earliest replication if multiple replications and if multiple replications all were failed (as per author descriptions)
```{r}
df <- df %>%
  group_by(ref_o) %>%
  fill(doi_o, .direction = "downup") %>%
  ungroup()

df <- df %>%
  group_by(ref_r) %>%
  fill(doi_r, .direction = "downup") %>%
  ungroup()

#check doi_r is not missing where it should not, fill in doi manually if missing
df %>% filter(is.na(doi_r))

#only originals with all failed replications
df <- df %>%
  mutate(r_key = if_else(is.na(doi_r), url_r, doi_r))

# identify doi_o where every replication attempt failed
all_failed_dois <- df %>%
  group_by(doi_o, r_key) %>%
  summarise(attempt_failed = all(pval_value_r > 0.05, na.rm = TRUE), .groups = "drop") %>%
  group_by(doi_o) %>%
  summarise(all_attempts_failed = all(attempt_failed), .groups = "drop") %>%
  filter(all_attempts_failed) %>%
  pull(doi_o)

df <- df %>%
  filter(doi_o %in% all_failed_dois) %>%
  select(-r_key)

```
# only failed replications
```{r}
# mutate dois to lower case
df <- df %>% mutate(doi_o = tolower(as.character(doi_o)))
df <- df %>% mutate(doi_r = tolower(as.character(doi_r)))
```

# getting metadata for the original studies
```{r}
# get dois for originals studies (here from Open Alex)
dois <- unique(na.omit(df$doi_o))
email <- Sys.getenv("MY_EMAIL")
##add email address to run!
options(openalexR.mailto = email)
# fetch metadata
OA_list <- oa_fetch(
  entity = "works",
  doi = dois,
  output = "tibble",
  verbose = TRUE
) 

# options = list(
#     select = c("id", "doi", "display_name", "publication_year", "primary_location.host_venue")
#   )


#test <- oa_fetch(
#   entity = "works",
#   doi = "10.1111/j.1468-0068.2010.00795.x",
#   output="list"
# ) 
meta_origOA <- OA_list
#meta_origOA <- fromJSON(toJSON(OA_list, auto_unbox = TRUE), flatten = TRUE) |> 
 # as_tibble()
##remove web address part of doi
meta_origOA <- meta_origOA %>%
  mutate(doi = stringr::str_remove(doi, regex("^https?://(dx\\.)?doi\\.org/", TRUE)))
df_meta <- dplyr::left_join(df, meta_origOA, by = c("doi_o" = "doi"))

```

# replication years
```{r}
doi_r <- unique(na.omit(df$doi_r))
##add email address to run!
options(openalexR.mailto = email)
# fetch metadata
OA_list_r <- oa_fetch(
  entity = "works",
  doi = doi_r,
  output = "tibble",
  verbose = TRUE
) 

saveRDS(OA_list_r,"data/metadata_r_OpenAlex.rds")

```
#append and select only earliest replication study
- replication studies without doi will not have a publication year on OpenAlex and these years will be added manually
```{r}
year_r <- OA_list_r %>%
  dplyr::select(
    doi_r = doi,
    publication_year_r = publication_year
  )
year_r <- year_r %>% mutate(doi_r = stringr::str_remove(doi_r, regex("^https?://(dx\\.)?doi\\.org/", TRUE)))
df_meta <- dplyr::left_join(df_meta, year_r, by = "doi_r")
colnames(df_meta)

df_meta <- df_meta %>%
  group_by(doi_o) %>%
  slice_min(publication_year_r, n = 1, with_ties = TRUE, na_rm = TRUE) %>%
  ungroup()


#select only studies which have replications publisjed between 2011 and 2021
df_meta <- df_meta %>% 
  filter(publication_year_r >= 2011 & publication_year_r <= 2021)
```

```{r}
#save the metadata
saveRDS(df_meta,file="data/metadata_OpenAlex.rds")

```





