---
title: "_05a_pre_analysis_data_prep"
author: "Josefina Weinerova"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# package loading
```{r}
if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  tidyverse, httr, jsonlite,
  purrr, dplyr, tibble, data.table,
  stringr, rscopus
)

source("R/api_cache.R")
source("R/analysis_prep_helpers.R")
```

#load data
```{r}
citations_controlGroup_Scopus <- readRDS("data/citing_studies_scopus.rds")
full_matched_df <- readRDS("data/full_matched_df.rds")
df_meta <- readRDS("data/metadata_OpenAlex.rds")
citing_studies_Scopus_rep <- readRDS("data/citing_studies_scopus_rep.rds")
```

#get year column from full date
```{r}
citations_controlGroup_Scopus <- citations_controlGroup_Scopus %>%
  mutate(year = as.integer(substr(citing_date, 1, 4)))
```

# get number of citations per year
```{r}
control_cit_per_year <- citations_controlGroup_Scopus %>%
  dplyr::group_by(doi_queried, year) %>%
  dplyr::summarise(n_citations = n(), .groups = "drop")
control_cit_per_year <- control_cit_per_year %>%
  filter(!is.na(year))
analysis_df_Scopus <- control_cit_per_year

null_doi <- setdiff(full_matched_df$doi, analysis_df_Scopus$doi_queried)
length(setdiff(analysis_df_Scopus$doi_queried,full_matched_df$doi))
null_doi <- null_doi[!is.na(null_doi) & null_doi != "NA" & null_doi != ""]
null_doi <- unique(null_doi)
length(null_doi)
```

#checking whether not returned dois are null
```{r}
# # this API endpoint seems to not work at the moment
# key <- Sys.getenv("SCOPUS_API_KEY")

# results <- purrr::map(
#   null_doi,
#   ~ cached_citation_retrieval(doi = .x, api_key = key, date_range = c("1900", "2025"))
# )
# ok <- purrr::map_lgl(results, ~ .x$status_code < 400)

# table(ok) 

# parsed <- results[ok] %>%
#   purrr::map("parsed")
# dois_ok <- null_doi[ok]

# str(parsed[[1]]$citation_info)
# names(parsed[[1]]$citation_info)

# citations_null <- purrr::imap_dfr(parsed, \(x, i) {
#   x$citation_info %>%
#     select(matches("^cc\\.\\d{4}$")) %>%
#     mutate(doi = dois_ok[[i]], .before = 1)
# }) %>%
#   pivot_longer(
#     cols = matches("^cc\\.\\d{4}$"),
#     names_to = "year",
#     values_to = "citations"
#   ) %>%
#   mutate(
#     year = as.integer(str_remove(year, "^cc\\.")),
#     citations = as.integer(citations)
#   )
```

#alternative API endpoint to get confirmation for whether not returned dois have 0 citations or are not indexed in Scopus
```{r}
rscopus::set_api_key(key)

results <- purrr::set_names(null_doi) %>%
  purrr::map(~ cached_scopus_cite_count(.x, api_key = key))


# which returned a clean response
ok <- purrr::map_lgl(results, ~ !is.na(.x$found))
table(ok)

# build summary table — one row per DOI
cite_check <- purrr::imap_dfr(results, ~ tibble(
  doi         = .y,
  status_code = .x$status_code,
  found       = .x$found,
  count       = .x$count
)) %>%
  mutate(verdict = case_when(
    is.na(found)  ~ "API error – inconclusive",
    !found        ~ "Not indexed in Scopus – _04a zero is an indexing gap",
    count == 0    ~ "Indexed, truly 0 citations",
    count  > 0    ~ "Indexed, has citations – possible _04a retrieval error"
  ))

table(cite_check$verdict)

#append true 0 doi
citations_null <- cite_check %>%
  filter(count == 0) %>%
  select(doi_queried = doi) %>%
  left_join(
    full_matched_df %>% select(doi_queried = doi, publication_year),
    by = "doi_queried"
  ) %>%
  filter(!is.na(publication_year)) %>%
  rowwise() %>%
  mutate(year = list(seq(publication_year, 2025))) %>%
  unnest(year) %>%
  mutate(n_citations = 0L) %>%
  select(doi_queried, year, n_citations)
```
```{r}
analysis_df_Scopus <- bind_rows(analysis_df_Scopus, citations_null)
```

```{r}
# key <- Sys.getenv("SCOPUS_API_KEY")

# # res <- rscopus::citation_retrieval(
# #   doi = "10.1038/ng.2429",
# #   api_key = key,
# #   date_range = c("1900", "2025")
# # )

# httr::status_code(res$get_statement)  # check if it worked
# rscopus::parse_citation_retrieval(res) # parse the result

# # other endpoint

# key <- Sys.getenv("SCOPUS_API_KEY")

# # Step 1: get the EID for your DOI
# res0 <- rscopus::scopus_search(
#   query = "DOI(10.1016/j.jfineco.2014.02.006)",
#   count = 1,
#   api_key = key
# )
# eid <- rscopus::gen_entries_to_df(res0$entries)$df$eid[1]

# # Step 2: get citing studies
# citers <- rscopus::scopus_search(
#   query = paste0("REFEID(", eid, ")"),
#   count = 200,
#   api_key = key
# )
```
#metadata for replications - year of publication
- ~~some replication studies do not have doi, and therefore cannot get metadata through API, the year will be filled manually from web search for those~~ [year is coded in FLoRA so not necessary anymore]

# pre-analysis modification

```{r}
replic_year <- df_meta %>% dplyr::select(doi_o, doi_r, publication_year_r)
orig_year <- full_matched_df %>% dplyr::select(doi, publication_year, issn_l)

analysis_df_Scopus <- prep_analysis_df(analysis_df_Scopus, replic_year, orig_year)
saveRDS(analysis_df_Scopus, "data/analysis_df_Scopus.rds")
```

# add years with 0 for years that are later than publication but have no data

```{r}
analysis_padded_Scopus <- pad_panel(analysis_df_Scopus)
```

#citation without co-citation

```{r}
citing_studies_Scopus_rep <- citing_studies_Scopus_rep %>% rename(doi_queried_r = doi_queried)
citing_studies_Scopus_rep <- citing_studies_Scopus_rep %>%
  left_join(
    replic_year,
    by = c("doi_queried_r" = "doi_r"),
    relationship = "many-to-many"
  )

analysis_padded_Scopus <- add_nococit(
  citations_raw = citations_controlGroup_Scopus,
  citations_rep = citing_studies_Scopus_rep,
  citing_id_col = "citing_eid",
  analysis_padded = analysis_padded_Scopus
)
```

#save the cleaned dataset
```{r}
saveRDS(analysis_padded_Scopus, "data/analysis_pad_Scopus.rds")
```
