---
title: "_05a_pre_analysis_data_prep"
author: "Josefina Weinerova"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# package loading
```{r}
if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  tidyverse, httr, jsonlite,
  purrr, dplyr, tibble, data.table,
  stringr, rscopus
)

source("R/api_cache.R")
source("R/analysis_prep_helpers.R")
```

#load data
```{r}
citations_controlGroup_Scopus <- readRDS("data/citing_studies_scopus.rds")
full_matched_df <- readRDS("data/full_matched_df.rds")
df_meta <- readRDS("data/metadata_OpenAlex.rds")
citing_studies_Scopus_rep <- readRDS("data/citing_studies_scopus_rep.rds")
```

#get year column from full date
```{r}
citations_controlGroup_Scopus <- citations_controlGroup_Scopus %>%
  mutate(year = as.integer(substr(citing_date, 1, 4)))
```

# get number of citations per year
```{r}
control_cit_per_year <- citations_controlGroup_Scopus %>%
  dplyr::group_by(doi_queried, year) %>%
  dplyr::summarise(n_citations = n(), .groups = "drop")
control_cit_per_year <- control_cit_per_year %>%
  filter(!is.na(year))
analysis_df_Scopus <- control_cit_per_year

null_doi <- setdiff(full_matched_df$doi, analysis_df_Scopus$doi_queried)
null_doi <- null_doi[!is.na(null_doi) & null_doi != "NA" & null_doi != ""]
null_doi <- unique(null_doi)
length(null_doi)
```

#checking whether not returned dois are null
```{r}
key <- Sys.getenv("SCOPUS_API_KEY")

results <- purrr::map(
  null_doi,
  ~ cached_citation_retrieval(doi = .x, api_key = key, date_range = c("1900", "2025"))
)
ok <- purrr::map_lgl(results, ~ .x$status_code < 400)

table(ok)

parsed <- results[ok] %>%
  purrr::map("parsed")
dois_ok <- null_doi[ok]

str(parsed[[1]]$citation_info)
names(parsed[[1]]$citation_info)

citations_null <- purrr::imap_dfr(parsed, \(x, i) {
  x$citation_info %>%
    select(matches("^cc\\.\\d{4}$")) %>%
    mutate(doi = dois_ok[[i]], .before = 1)
}) %>%
  pivot_longer(
    cols = matches("^cc\\.\\d{4}$"),
    names_to = "year",
    values_to = "citations"
  ) %>%
  mutate(
    year = as.integer(str_remove(year, "^cc\\.")),
    citations = as.integer(citations)
  )

citations_null <- citations_null %>%
  dplyr::group_by(doi) %>%
  filter(all(citations == 0))
colnames(citations_null) <- colnames(analysis_df_Scopus)

analysis_df_Scopus <- bind_rows(analysis_df_Scopus, citations_null)
```


#metadata for replications - year of publication
- some replication studies do not have doi, and therefore cannot get metadata through API, the year will be filled manually from web search for those

# pre-analysis modification

```{r}
replic_year <- df_meta %>% dplyr::select(doi_o, doi_r, publication_year_r)
orig_year <- full_matched_df %>% dplyr::select(doi, publication_year, issn_l)

analysis_df_Scopus <- prep_analysis_df(analysis_df_Scopus, replic_year, orig_year)
saveRDS(analysis_df_Scopus, "data/analysis_df_Scopus.rds")
```

# add years with 0 for years that are later than publication but have no data

```{r}
analysis_padded_Scopus <- pad_panel(analysis_df_Scopus)
```

#citation without co-citation

```{r}
citing_studies_Scopus_rep <- citing_studies_Scopus_rep %>% rename(doi_queried_r = doi_queried)
citing_studies_Scopus_rep <- citing_studies_Scopus_rep %>%
  left_join(
    replic_year,
    by = c("doi_queried_r" = "doi_r"),
    relationship = "many-to-many"
  )

analysis_padded_Scopus <- add_nococit(
  citations_raw = citations_controlGroup_Scopus,
  citations_rep = citing_studies_Scopus_rep,
  citing_id_col = "citing_eid",
  analysis_padded = analysis_padded_Scopus
)
```

#save the cleaned dataset
```{r}
saveRDS(analysis_padded_Scopus, "data/analysis_pad_Scopus.rds")
```
