---
title: "_05_preanalysis_data_prep"
author: "Josefina Weinerova"
date: "2025-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# package loading
```{r}
if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  openalexR, tidyverse, httr, jsonlite,
  purrr, dplyr, tibble, data.table
)

source("R/api_cache.R")
source("R/analysis_prep_helpers.R")
```

#load data
```{r}
citations_controlGroup <- readRDS("data/citations_OC.rds")
full_matched_df <- readRDS("data/full_matched_df.rds")
df_meta <- readRDS("data/metadata_OpenAlex.rds")
citations_rep <- readRDS("data/citations_OC_rep.rds")
```

# get number of citations per year
```{r}
control_cit_per_year <- citations_controlGroup %>%
  dplyr::group_by(doi_queried, year) %>%
  dplyr::summarise(n_citations = n(), .groups = "drop")
control_cit_per_year <- control_cit_per_year %>%
  filter(!is.na(year))
analysis_df <- control_cit_per_year

null_doi <- setdiff(full_matched_df$doi, analysis_df$doi_queried)
null_doi <- null_doi[!is.na(null_doi) & null_doi != "NA" & null_doi != ""]
null_doi <- unique(null_doi)
```

#do missing dois have 0 citations
```{r}
email = Sys.getenv("MY_EMAIL")
options(openalexR.mailto = email)
batches <- split(null_doi, ceiling(seq_along(null_doi) / 50))

safe_fetch <- purrr::safely(function(d) {
  cached_oa_fetch(
    entity = "works",
    doi = d,
    options = list(select = c("doi", "cited_by_count", "counts_by_year")),
    output = "tibble",
    mailto = email,
    cache_key = list(stage = "openalex_zero_citation_check"),
    verbose = TRUE
  )
})

res <- purrr::imap(batches, \(d, i) {
  Sys.sleep(0.2)
  out <- safe_fetch(d)
  list(
    batch = i,
    n_dois = length(d),
    data = out$result %||% tibble(),
    error = if (!is.null(out$error)) conditionMessage(out$error) else NA_character_
  )
})

null_cit <- purrr::map_dfr(res, "data")
errors <- dplyr::bind_rows(purrr::map(res, \(x) tibble(batch = x$batch, n_dois = x$n_dois, error = x$error))) %>%
  dplyr::filter(!is.na(error))

true_null <- null_cit %>% filter(cited_by_count == 0)
true_null <- true_null %>%
  mutate(
    doi = doi %>%
      as.character() %>%
      str_trim() %>%
      str_replace("^https?://(dx\\.)?doi\\.org/", "") %>%
      str_replace("^doi:\\s*", "") %>%
      tolower()
  )

matched_years <- full_matched_df %>%
  transmute(doi, publication_year) %>%
  distinct(doi, .keep_all = TRUE)

true_null <- true_null %>%
  left_join(matched_years, by = "doi")

true_null <- true_null %>%
  mutate(publication_year = as.integer(publication_year)) %>%
  rowwise() %>%
  mutate(
    year = list(
      if (is.na(publication_year)) {
        NA_integer_
      } else if (publication_year <= 2025L) {
        seq.int(publication_year, 2025L)
      } else {
        publication_year
      }
    )
  ) %>%
  ungroup() %>%
  tidyr::unnest(year)
true_null <- true_null %>%
  mutate(n_citations = cited_by_count, doi_queried = doi) %>%
  dplyr::select(doi_queried, n_citations, year)
analysis_df <- bind_rows(analysis_df, true_null)
```

#metadata for replications - year of publication
- some replication studies do not have doi, and therefore cannot get metadata through API, the year will be filled manually from web search for those

# pre-analysis modification

```{r}
replic_year <- df_meta %>% dplyr::select(doi_o, doi_r, publication_year_r)
orig_year <- full_matched_df %>% dplyr::select(doi, publication_year, issn_l)

analysis_df <- prep_analysis_df(analysis_df, replic_year, orig_year)
saveRDS(analysis_df, "data/analysis_df.rds")
```

# add years with 0 for years that are later than publication but have no data

```{r}
analysis_padded <- pad_panel(analysis_df)
```

## Citations without co-citation
```{r}
citations_rep <- citations_rep %>% rename(doi_queried_r = doi_queried)
citations_rep <- citations_rep %>%
  left_join(
    replic_year,
    by = c("doi_queried_r" = "doi_r"),
    relationship = "many-to-many"
  )

analysis_padded <- add_nococit(
  citations_raw = citations_controlGroup,
  citations_rep = citations_rep,
  citing_id_col = "citing",
  analysis_padded = analysis_padded
)
```

#save the cleaned dataset
```{r}
saveRDS(analysis_padded, "data/analysis_pad.rds")
```
