---
title: "_06_fect_analysis"
author: "Josefina Weinerova"
date: "2025-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# package loading
```{r}
if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  openalexR, cowplot, tidyverse, httr, jsonlite,
  citecorp, rcrossref, purrr, dplyr, tibble, fect, furrr, future, data.table,
  panelView, readxl, fixest, etwfe
)
```

# analysis - load analysis_pad file here if starting from scratch
```{r}
df_journals_L <- readRDS("data/df_journals_L.rds")
```

#switch data source
```{r}
# Choose data source: "OC" or "Scopus"
data_source <- "OC"

analysis_padded <- switch(data_source,
  "OC" = readRDS("data/analysis_pad.rds"),
  "Scopus" = readRDS("data/analysis_pad_Scopus.rds"),
  stop("data_source must be either 'OC' or 'Scopus'")
)
```

## Sample description
Check if control to treated numbers fit

```{r}
analysis_padded %>%
  group_by(doi_queried) %>%
  summarise(
    ever_treated = any(D == 1),
    treat_year = ifelse(ever_treated, min(year[D == 1]), NA_integer_),
    pre_n = sum(year < treat_year & (D == 0 | !ever_treated)),
    post_n = sum(year >= treat_year & is.finite(treat_year)),
    missing_pub_year = all(is.na(publication_year)),
    missing_pub_year_any = any(is.na(publication_year)),
    inconsistent_years = length(na.omit(unique(publication_year_r))) > 1
  ) %>%
  summarise(
    n_units = n(),
    n_treated = sum(ever_treated),
    n_control = n_units - n_treated,
    pre_n_median = median(pre_n[ever_treated], na.rm = TRUE),
    pre_n_min = min(pre_n[ever_treated], na.rm = TRUE),
    pre_n_max = max(pre_n[ever_treated], na.rm = TRUE),
    pre_n_below_5 = sum(pre_n[ever_treated] < 5, na.rm = TRUE),
    post_n_median = median(post_n[ever_treated], na.rm = TRUE),
    post_n_min = min(post_n[ever_treated], na.rm = TRUE),
    post_n_max = max(post_n[ever_treated], na.rm = TRUE),
    n_missing_pub_year = sum(missing_pub_year),
    n_missing_pub_year_any = sum(missing_pub_year_any),
    n_inconsistent_years = sum(inconsistent_years)
  ) %>%
  pivot_longer(everything(), names_to = "stat", values_to = "value")


analysis_padded %>%
  select(doi_queried, publication_year, D) %>%
  group_by(doi_queried, publication_year) %>%
  summarise(ever_treated = any(D == 1)) %>%
  ungroup() %>%
  ggplot(aes(x = publication_year, fill = as.factor(ever_treated))) +
  geom_bar(position = "dodge")

analysis_padded %>%
  select(doi_queried, publication_year, D) %>%
  group_by(doi_queried, publication_year) %>%
  dplyr::summarise(ever_treated = any(D == 1)) %>%
  ungroup() %>%
  count(publication_year, ever_treated) %>%
  pivot_wider(names_from = ever_treated, values_from = n, values_fill = 0) %>%
  mutate(ratio = `FALSE` / `TRUE`)


analysis_padded %>%
  dplyr::group_by(doi_queried) %>%
  filter(any(D == 1)) %>%
  ungroup() %>%
  mutate(unit_order = match(doi_queried, unique(doi_queried))) %>%
  panelview(n_citations ~ D,
    index = c("unit_order", "year"),
    xlab = "Year", ylab = "Unit", display.all = T,
    gridOff = TRUE
  )


analysis_padded %>% head()
```
# add covariates
```{r}
analysis_padded <- analysis_padded %>% mutate(issn_l = str_replace_all(issn_l, "-", ""))
# sjr
sjr <- df_journals_L %>%
  mutate(issn_l = issn) %>%
  select(year, issn_l, sjr)
analysis_padded <- merge(
  x = analysis_padded,
  y = sjr,
  by = c("issn_l", "year"),
  all.x = TRUE # keep all rows from analysis_padded (left join)
)
# age2
analysis_padded <- analysis_padded %>%
  group_by(doi_queried) %>%
  mutate(
    age  = year - publication_year,
    age2 = age^2
  ) %>%
  ungroup() %>%
  arrange(doi_queried, year)

# forward fill sjr
analysis_padded <- analysis_padded %>%
  arrange(doi_queried, year) %>%
  group_by(doi_queried) %>%
  fill(sjr, .direction = "down") %>%
  ungroup()

analysis_padded <- analysis_padded %>% filter(!is.na(sjr))
```

# analysis 
```{r}
str(analysis_padded)

# Identify number of factors

fit_factor <- fect(
  n_citations ~ D + age2 + sjr,
  data = analysis_padded,
  index = c("doi_queried", "year"),
  method = "ife",
  force = "two-way",
  se = TRUE,
  CV = TRUE,
  r = 0:12,
  nboots = 500,
  keep.sims = TRUE,
  min.T0 = 5
)
```


# analysis - run only with the factor selected as best
```{r}
fit <- fect(
  n_citations ~ D + age2 + sjr,
  data = analysis_padded,
  index = c("doi_queried", "year"),
  method = "ife",
  force = "two-way",
  se = TRUE,
  CV = TRUE,
  r = 2,
  nboots = 500,
  keep.sims = TRUE,
  min.T0 = 3
)


print(fit)

plot(fit,
  main = "Estimated treatment effect", ylab = "Effect of failed replication on citations",
  cex.main = 0.8, cex.lab = 0.8, cex.axis = 0.8
)


saveRDS(fit, paste0("models/fit_citation_", data_source, ".rds"))
```

# control tests
```{r}
# placebo - checks for pre-treatment effects
placebo_win <- c(-4, -1)

fit_placebo <- fect(
  n_citations ~ D + age2 + sjr,
  data = analysis_padded,
  index = c("doi_queried", "year"),
  method = "ife",
  force = "two-way",
  r = 2,
  se = TRUE,
  CV = FALSE,
  nboots = 200,
  placeboTest = TRUE,
  placebo.period = placebo_win,
  min.T0 = 3
)

fit_placebo$test.out
p.out <- plot(
  fit_placebo,
  type = "equiv", start0 = TRUE,
  highlight = TRUE,
  stats = c("placebo.p", "equiv.p"),
  return.test = TRUE
)
p.out # contains placebo test statistics/p-values
```

# pre trend F test on the main model
- evaluate pre treatment fit
```{r}
res <- plot(
  fit,
  type = "gap",
  start0 = TRUE,
  return.test = TRUE,
  stats = "F.p"
)

res$test.out
```



## citation without co citation

# analysis 
```{r}
str(analysis_padded)

fit_factor_nococit <- fect(
  n_nococit ~ D + age2 + sjr,
  data = analysis_padded,
  index = c("doi_queried", "year"),
  method = "ife",
  force = "two-way",
  se = TRUE,
  CV = TRUE,
  r = 1:4,
  nboots = 500,
  keep.sims = TRUE,
  min.T0 = 5
)
```

# analysis - run only with the factor selected as best
```{r}
fit_nococit <- fect(
  n_nococit ~ D + age2 + sjr,
  data = analysis_padded,
  index = c("doi_queried", "year"),
  method = "ife",
  force = "two-way",
  se = TRUE,
  CV = TRUE,
  r = 2,
  nboots = 300,
  keep.sims = TRUE,
  min.T0 = 3
)


print(fit_nococit)

plot(fit_nococit,
  main = "Estimated average treatment effect", ylab = "Effect of failed replication on citations without co-citation",
  cex.main = 0.8, cex.lab = 0.8, cex.axis = 0.8
)


saveRDS(fit_nococit, paste0("models/fit_nococitation_", data_source, ".rds"))
```
#control tests
```{r}
# placebo - checks for pre-treatment effects
placebo_win <- c(-4, -1)

fit_placebo <- fect(
  n_nococit ~ D + age2 + sjr,
  data = analysis_padded,
  index = c("doi_queried", "year"),
  method = "ife",
  force = "two-way",
  r = fit_factor_nococit$r.cv,
  se = TRUE,
  CV = FALSE,
  nboots = 200,
  placeboTest = TRUE,
  placebo.period = placebo_win,
  min.T0 = 3
)

fit_placebo$test.out
p.out <- plot(
  fit_placebo,
  type = "equiv", start0 = TRUE,
  highlight = TRUE,
  stats = c("placebo.p", "equiv.p"),
  return.test = TRUE
)
p.out # contains placebo test statistics/p-values
```

# pre trend F test on the main model
- evaluate pre treatment fit
```{r}
res <- plot(
  fit_nococit,
  type = "gap",
  start0 = TRUE,
  return.test = TRUE,
  stats = "F.p"
)

res$test.out
```


#robustness check with etwfe
```{r}
# Step 1: Derive treatment timing variable
# This is the first year D == 1 for each study
analysis_padded <- analysis_padded %>%
  dplyr::group_by(doi_queried) %>%
  mutate(
    g = if_else(any(D == 1), min(year[D == 1], na.rm = TRUE), 0L)
  ) %>%
  ungroup()

analysis_etwfe <- analysis_padded %>%
  filter(
    !is.na(sjr)
  )

# model
# Set fixest options globally before fitting
setFixest_estimation(glm.iter = 10000, fixef.iter = 10000)
etwfe_fit <- etwfe(
  fml       = n_citations ~ sjr + age2, # controls; unit & time FE added automatically
  tvar      = year, # time variable
  gvar      = g, # treatment timing (0 = never treated)
  data      = analysis_etwfe,
  ivar      = doi_queried, # unit identifier
  family    = "poisson",
  vcov      = ~doi_queried
)

summary(etwfe_fit)

iplot(etwfe_fit)

# Step 3: Aggregate to a single ATT
att <- emfx(etwfe_fit, type = "simple")
print(att)

# Step 4: Dynamic (event-study) plot — equivalent to fect's ATT over time
att_dynamic <- emfx(etwfe_fit, type = "event")
print(att_dynamic)

# Plot the event study
att_dynamic %>%
  ggplot(aes(
    x = event, y = estimate,
    ymin = conf.low, ymax = conf.high
  )) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(alpha = 0.2, fill = "steelblue") +
  geom_line(colour = "steelblue") +
  geom_point(colour = "steelblue") +
  labs(
    x = "Years relative to failed replication",
    y = "ATT (citation count)",
    title = "Effect of failed replication on citation counts",
    subtitle = "ETWFE with Poisson family"
  ) +
  theme_minimal()

emfx(etwfe_fit, type = "event") |> plot() 
```

# for citations without co_citation
```{r}
# model for citation without co-citation
# Set fixest options globally before fitting
setFixest_estimation(glm.iter = 10000, fixef.iter = 10000)
etwfe_fit_nococit <- etwfe(
  fml       = n_nococit ~ sjr + age2, # controls; unit & time FE added automatically
  tvar      = year, # time variable
  gvar      = g, # treatment timing (0 = never treated)
  data      = analysis_etwfe,
  ivar      = doi_queried, # unit identifier
  family    = "poisson",
  vcov      = ~doi_queried
)

summary(etwfe_fit_nococit)

iplot(etwfe_fit_nococit)

# Step 3: Aggregate to a single ATT
att <- emfx(etwfe_fit_nococit, type = "simple")
print(att)

# Step 4: Dynamic (event-study) plot — equivalent to fect's ATT over time
att_dynamic <- emfx(etwfe_fit_nococit, type = "event")
print(att_dynamic)

# Plot the event study
att_dynamic %>%
  ggplot(aes(
    x = event, y = estimate,
    ymin = conf.low, ymax = conf.high
  )) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(alpha = 0.2, fill = "steelblue") +
  geom_line(colour = "steelblue") +
  geom_point(colour = "steelblue") +
  labs(
    x = "Years relative to failed replication",
    y = "ATT (citation count)",
    title = "Effect of failed replication on citation without co-citation counts",
    subtitle = "ETWFE with Poisson family"
  ) +
  theme_minimal()
```


