---
title: "Successful versus failed replications: citation effects"
output: html_document
params:
  result_path: "results/replication_contrast/result.rds"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
if (!file.exists(params$result_path)) {
  stop("Run R/prepare_replication_contrast_inputs.R and R/run_replication_contrast.R first; see docs/replication_contrast.md.")
}
result <- readRDS(params$result_path)
library(ggplot2)
```

The contrast is the average citation effect following a successful replication
minus that following a failed replication. Each effect is relative to the
estimated citation trajectory without a recorded replication. Both groups use
the same replication-cohort weights and event-time window. These are effects
in the respective outcome groups; their difference does not identify what
would happen if the same replication succeeded instead of failing.

```{r inference-status, results='asis'}
if (!result$inference_ok) cat("**Bootstrap inference failed. Confidence intervals are withheld.**\n")
cat("Selected latent factors:", result$settings$selected_r,
    ". Requested bootstrap draws:", result$settings$nboots, ".\n")
cat("Primary event years:", paste(result$settings$primary_times, collapse = ", "), ".\n")
```

```{r primary}
knitr::kable(subset(result$estimates, period == "primary"), digits = 3)
```

```{r event-study, fig.width=9, fig.height=4.5}
curves <- subset(result$estimates, period != "primary")
curves$event_time <- as.integer(curves$period)
ggplot(curves, aes(event_time, estimate)) +
  geom_hline(yintercept = 0, colour = "grey65") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = .18, na.rm = TRUE) +
  geom_line(colour = "steelblue4") + geom_point(colour = "steelblue4") +
  facet_wrap(~estimand) + theme_minimal() +
  labs(x = "Years since first recorded replication", y = "Citations per paper-year")
```

Intervals use the standard deviation of a common cluster bootstrap. Each draw
refits the IFE model and calculates both effects and their difference. The
factor count and target cohort weights are held fixed after selection on the
original sample. Intervals are pointwise, not simultaneous event-study bands.

The target sample has observations throughout the event window. Follow-up ends
before the year of a later recorded attempt. Ambiguous first outcomes, undated
histories, and first events outside the replication/cohort definition are
excluded from both treatment groups and the untreated controls.

```{r support}
knitr::kable(result$support)
knitr::kable(data.frame(cohort = names(result$weights), weight = unname(result$weights)), digits = 3)
knitr::kable(as.data.frame(with(result$audit,
  table(replication_outcome, in_panel, in_fit, in_contrast))))
```

Pretreatment prediction diagnostics hide the two years immediately before
publication and refit the model. The table shows held-out citation gaps in
each group and their difference. These point diagnostics cannot establish that
post-treatment counterfactual predictions are valid. An error indicates that
the selected sample/model could not support the diagnostic.

```{r placebo}
knitr::kable(result$placebo, digits = 3)
```

```{r bootstrap-failures}
knitr::kable(subset(result$failures, !is.na(error)))
```
