---
title: "_0_power"
author: "Josefina Weinerova"
date: "2025-11-21"
output: html_document
---
```{r}
# install.packages("openalexR")
if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  openalexR, cowplot, tidyverse, httr, jsonlite,
  citecorp, rcrossref, purrr, dplyr, tibble, fect, furrr, future, data.table,
  panelView, rscopus, httr2, plotly, MASS, GGally, metafor, progressr, readxl, openxlsx
)

# loading data
analysis_df <- readxl::read_excel("data/analysis_df.xlsx")
cos_sample <- readxl::read_excel("2025-02-07_COSdata_validated_delivered.xlsx")
all_matched_articles <- readxl::read_excel("data/all_matched_articles_pilot.xlsx")


plan(multisession, workers = availableCores() - 1) 

my_seed <- 123
```


## Power
## Data-Generating Process (Simulation Outline)

1. Get summary metrics from a sample of control studies
- first year citations
- cumulative last year citations
- number of years
- average growth (average number of citations per year)
- early growth (average number of citations in first 3 years)

2. Create Gaussian multivariate distribution by first ranking the individual metrics and then creating normal distribution
- calculate correlation among different summary metrics
- in the normal space simulate 100,000 new studies by sampling from the distribution while keeping the correlation intact
- transport back to original space by inverse function

# @Josefina: clarify above that you did not simulate collinear variables (e.g., total citations and average growth are collinear)

3. Use the 5 summary metrics to create citation trajectories of the simulated studies

4. select 10,000 studies as treated studies
- select year of treatment start; effect starts 1 year later (use `D_lag`)
- apply effect via *Poisson thinning*: for treated-years, `y_new ~ Binomial(y, p = effect_mult)`, which preserves the discrete observed trajectory while reducing expected counts

5. run fect analysis 100 times for each effect size, collect p.values and ATTs

# what do real citation trajectories look like   

```{r}
analysis_df$year <- as.numeric(analysis_df$year)
treated_dois <- cos_sample$doi_o
analysis_df <- analysis_df %>%
  dplyr::group_by(doi_queried) %>%
  complete(year = full_seq(c(min(year), max(year)), 1), fill = list(n_citations = 0)) %>%
  ungroup()

doi_sample <- analysis_df %>%
  dplyr::distinct(doi_queried) %>%
  dplyr::filter(!(doi_queried %in% treated_dois)) %>%
  dplyr::pull(doi_queried)

doi_sample <- sample(doi_sample, size = min(10000, length(doi_sample)))

# 1) Compute cumulative citations per doi over time for a sample of 200 dois
analysis_df_cum <- analysis_df %>% filter(doi_queried %in% doi_sample)
range(analysis_df_cum$year)
# checking for weird start years
analysis_df_cum <- analysis_df_cum %>%
  left_join(
    all_matched_articles %>%
      dplyr::select(doi, publication_year),
    by = c("doi_queried" = "doi")
  )

analysis_df_cum <- analysis_df_cum %>% filter(year >= publication_year)
analysis_df_cum <- analysis_df_cum %>%
  arrange(doi_queried, year) %>%
  dplyr::group_by(doi_queried) %>%
  mutate(cum_citations = cumsum(dplyr::coalesce(n_citations, 0))) %>%
  ungroup()

analysis_df_cum$year <- as.numeric(analysis_df_cum$year)
analysis_df_cum <- analysis_df_cum %>%
  dplyr::group_by(doi_queried) %>%
  mutate(time_since_pub = year - publication_year + 1) %>%
  ungroup()
summary(analysis_df_cum$cum_citations)
# not normalized
ggplot(
  analysis_df_cum,
  aes(
    x = time_since_pub,
    y = cum_citations,
    group = doi_queried
  )
) +
  geom_line(alpha = 0.1) +
  labs(
    x = "Year",
    y = "Cumulative citations",
    color = "DOI",
    title = "Cumulative citation growth over time"
  ) +
  theme_minimal() +
  scale_y_log10()

```
#plotting distribution of parameters

```{r}
citation_summary <- analysis_df_cum %>%
  dplyr::group_by(doi_queried) %>%
  arrange(time_since_pub) %>%
  mutate(
    yearly_cites = cum_citations - lag(cum_citations, default = 0)
  ) %>%
  dplyr::summarise(
    first_year_citations = yearly_cites[time_since_pub == 1],
    max_cumulative = max(cum_citations),
    years_observed = max(time_since_pub),
    avg_yearly_growth = mean(yearly_cites),
    early_growth = mean(yearly_cites[time_since_pub <= min(3, years_observed)]),
    linear_slope = coef(lm(cum_citations ~ time_since_pub))[2],
    log_slope = coef(lm(log1p(cum_citations) ~ time_since_pub))[2]
  )

citation_summary <- na.omit(citation_summary)
plot_ly(
  citation_summary,
  x = ~first_year_citations,
  y = ~max_cumulative,
  z = ~years_observed,
  color = ~doi_queried, # color by DOI
  colors = "Dark2",
  type = "scatter3d",
  mode = "markers"
)

citation_summary <- as.data.frame(citation_summary)

summary(citation_summary)
cor(citation_summary %>% dplyr::select(-doi_queried))


ggpairs(citation_summary, columns = 2:8)
```


#sampling from real distributions
```{r}
vars <- citation_summary %>%
  dplyr::select(
    first_year_citations,
    max_cumulative,
    years_observed,
    avg_yearly_growth,
    early_growth
  )

n <- nrow(vars)

# empirical CDF via ranks, then map to normal
to_normal <- function(x) {
  u <- (rank(x, ties.method = "average") - 0.5) / n # in (0,1)
  qnorm(u) # ≈ N(0,1)
}

Z <- vars %>%
  mutate(across(everything(), to_normal)) %>%
  as.matrix()


Sigma <- cor(Z)

n_sim <- 100000

Z_sim <- mvrnorm(n = n_sim, mu = rep(0, ncol(Z)), Sigma = Sigma)
colnames(Z_sim) <- colnames(vars)


# helper: empirical quantile function
inv_ecdf <- function(x, u) {
  # u in (0,1); use type=8 interpolation
  quantile(x, probs = u, type = 8, na.rm = TRUE)
}

U_sim <- pnorm(Z_sim) # now values in (0,1)

sim_vars <- matrix(NA, nrow = n_sim, ncol = ncol(vars))
colnames(sim_vars) <- colnames(vars)

for (j in seq_along(vars)) {
  sim_vars[, j] <- inv_ecdf(vars[[j]], U_sim[, j])
}

sim_summaries <- as.data.frame(sim_vars)

summary(sim_summaries)
```


#simulate from summaries
```{r}
simulate_yearly_from_summary <- function(
  Time,
  total,
  first1,
  early_growth = NULL,
  last_year = 2025
) {
  # Edge cases:
  Time <- max(1, round(Time))
  total <- max(0, round(total))
  first1 <- round(first1)

  # If no citations at all
  if (total == 0) {
    years <- seq(last_year - Time + 1, last_year)
    return(tibble(
      year = years,
      yearly_cites = rep(0, Time),
      cum_cites = rep(0, Time)
    ))
  }

  # If only 1 year observed, everything happens in that year
  if (Time == 1) {
    years <- last_year
    return(tibble(
      year = years,
      yearly_cites = total,
      cum_cites = total
    ))
  }

  # Ensure first-year citations are between 0 and total
  first1 <- max(0, min(first1, total))

  # ---- Baseline linear shape ----
  # yearly(t) = a + b*t
  b <- 2 * (total - Time * first1) / (Time * (Time - 1))
  a <- first1 - b

  t_seq <- 1:Time
  lam <- a + b * t_seq

  # ---- Shape adjustment using ONLY early_growth ----
  # derive the implied average yearly citations from (total, T)
  avg_growth_derived <- total / Time

  if (!is.null(early_growth) && is.finite(early_growth) && avg_growth_derived > 0) {
    # early_ratio > 1 -> front-loaded, < 1 -> back-loaded
    early_ratio <- early_growth / avg_growth_derived

    # clamp to avoid crazy numbers
    early_ratio <- pmin(pmax(early_ratio, 0.3), 3)

    # smooth warping parameter
    k <- log(early_ratio) / max(Time - 1, 1)

    # weights over time
    w <- exp(-k * (t_seq - 1))

    lam <- lam * w
  }

  # Enforce non-negativity
  lam[lam < 0] <- 0

  # If everything went to zero, fall back to uniform
  if (sum(lam) == 0) {
    lam <- rep(total / Time, Time)
  } else {
    # Rescale lambdas so they sum exactly to total in expectation
    lam <- lam * (total / sum(lam))
  }

  # Generate integer yearly citations (Poisson noise)
  yearly <- rpois(Time, lambda = lam)

  # Avoid the degenerate all-zero case
  if (sum(yearly) == 0) {
    yearly <- round(lam)
  }

  # Adjust mismatch so sum(yearly) == total
  diff <- total - sum(yearly)
  if (diff != 0) {
    idx <- sample(1:Time, size = abs(diff), replace = TRUE)
    for (i in idx) {
      yearly[i] <- yearly[i] + sign(diff)
    }
    yearly[yearly < 0] <- 0
  }

  cum <- cumsum(yearly)
  years <- seq(last_year - Time + 1, last_year)

  tibble(
    year = years,
    yearly_cites = yearly,
    cum_cites = cum
  )
}
```

#apply to simulated studies
```{r}
sim_summaries <- sim_summaries %>%
  mutate(doi_queried = row_number())


handlers(handler_rstudio())

with_progress({
  simulated_trajectories <- sim_summaries %>%
    dplyr::select(
      doi_queried,
      years_observed,
      max_cumulative,
      first_year_citations,
      early_growth
    ) %>%
    future_pmap_dfr(
      function(doi_queried,
               years_observed,
               max_cumulative,
               first_year_citations,
               early_growth) {
        simulate_yearly_from_summary(
          Time = years_observed,
          total = max_cumulative,
          first1 = first_year_citations,
          early_growth = early_growth,
          last_year = 2025
        ) %>%
          mutate(study_id = doi_queried, .before = 1)
      },
      .progress = TRUE,
      .options = furrr_options(seed = my_seed)
    )
})

range(simulated_trajectories$year)
simulated_trajectories <- simulated_trajectories %>%
  dplyr::group_by(study_id) %>%
  mutate(
    start_year = min(year),
    time_since_pub = year - start_year + 1
  ) %>%
  ungroup()

# not normalized
ggplot(
  simulated_trajectories,
  aes(
    x = time_since_pub,
    y = cum_cites,
    group = study_id
  )
) +
  geom_line(alpha = 0.1) +
  labs(
    x = "Year",
    y = "Cumulative citations",
    color = "DOI",
    title = "Simulated cumulative citation growth over time"
  ) +
  theme_minimal() +
  scale_y_log10()


```

```{r}

study_info <- simulated_trajectories %>%
  dplyr::group_by(study_id) %>%
  dplyr::summarise(
    start_year = min(year),
    end_year = max(year),
    .groups = "drop"
  )


study_info <- study_info %>%
  mutate(
    treat_from = pmax(start_year + 5, 2010L),
    treat_to   = 2024L
  ) %>%
  filter(treat_from <= treat_to)


summary(study_info)

# sample treated studies
set.seed(123)

treated_ids <- sample(study_info$study_id, size = 10000)

# assign treatment
treat_info <- study_info %>%
  filter(study_id %in% treated_ids) %>%
  rowwise() %>%
  mutate(
    treat_year = if_else(
      treat_from == treat_to,
      treat_from,
      sample(treat_from:treat_to, size = 1)
    )
  ) %>%
  ungroup() %>%
  dplyr::select(study_id, treat_year)


sim_with_treat <- simulated_trajectories %>%
  left_join(treat_info, by = "study_id") %>%
  mutate(
    treated = !is.na(treat_year),
    # D = 1 in the treatment year and all years after
    D = if_else(
      is.na(treat_year), # untreated studies
      0L,
      if_else(year >= treat_year, 1L, 0L)
    )
  )

sim_with_treat <- sim_with_treat %>%
  dplyr::group_by(study_id) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    D_lag = dplyr::lag(D, default = 0L) # D_{t-1}
  ) %>%
  ungroup()

## commented for the full run, this part is already implemented below, here it was used as checks for the effect implementation before the run

## Previous approach (stochastic rounding) kept for reference; main simulation now uses Poisson thinning
# sim_with_treat <- sim_with_treat %>%
#   mutate(
#     lambda_baseline = pmax(yearly_cites, 0),
#     lambda_treated  = if_else(
#       D_lag == 1L,                 # or D == 1L if you want instantaneous effect
#       0.8 * lambda_baseline,
#       lambda_baseline
#     ),
#     lambda_treated = pmax(lambda_treated, 0),   # safety
#
#     # stochastic rounding:
#     floor_val = floor(lambda_treated),
#     frac      = lambda_treated - floor_val,     # e.g. 0.80 for 3.80
#     frac      = pmin(pmax(frac, 0), 1),         # clamp to [0,1]
#
#     # add Bernoulli(frac): 1 with prob = frac, 0 otherwise
#     yearly_cites_new = floor_val + rbinom(n(), size = 1, prob = frac)
#   )
```
#plot after adding treatment
```{r}
set.seed(123)
example_effect_mult <- 0.9
sim_with_treat_example <- sim_with_treat %>%
  mutate(
    yearly_cites_new = if_else(
      D_lag == 1L,
      rbinom(n(), size = pmax(yearly_cites, 0), prob = example_effect_mult), 
      yearly_cites 
    )
  ) %>%
  dplyr::group_by(study_id) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(cum_cites_new = cumsum(yearly_cites_new)) %>%
  ungroup()


sim_with_treat_example <- sim_with_treat_example %>%
  group_by(study_id) %>%
  mutate(
    publication_year = min(year),
    time_since_pub = year - publication_year + 1
  )

ggplot(
  sim_with_treat_example,
  aes(
    x = time_since_pub,
    y = cum_cites_new,
    group = study_id, colour = factor(D_lag)
  )
) +
  geom_line(alpha = 0.2) +
  labs(
    x = "Year",
    y = "Cumulative citations",
    color = "Treatment status", # @Josefina: renamed from DOI - right?
    title = "Simulated cumulative citation growth over time"
  ) +
  theme_minimal() +
  scale_y_log10()

a <- sim_with_treat_example %>%
  dplyr::group_by(study_id) %>%
  dplyr::filter(any(D == 1L)) %>%
  dplyr::filter(study_id == 5)

ggplot(a, aes(x = year, group = study_id)) +
  geom_line(aes(y = yearly_cites, colour = "baseline")) +
  geom_line(aes(y = yearly_cites_new, colour = "thinned")) +
  labs(y = "Yearly citations", colour = NULL)

postlag_by_id <- sim_with_treat %>%
  filter(treated) %>%
  group_by(study_id) %>%
  summarise(
    end_year = max(year),
    treat_year = first(treat_year),
    postlag_years = sum(D_lag == 1L),
    .groups = "drop"
  )

summary(postlag_by_id$postlag_years)
```


```{r}
unit_treat_status <- sim_with_treat %>%
  dplyr::group_by(study_id) %>%
  dplyr::summarise(
    ever_treated = any(D == 1L),
    .groups = "drop"
  )

treated_units <- unit_treat_status$study_id[unit_treat_status$ever_treated]
untreated_units <- unit_treat_status$study_id[!unit_treat_status$ever_treated]
```

```{r}
run_one_fect_sim <- function(iter,
                             sim_with_treat,
                             treated_units,
                             untreated_units,
                             n_treated = 1000,
                             n_untreated = 10000,
                             effect_mult = 0.9,
                             parallel_fect = TRUE) {
  set.seed(iter) # optional: reproducible per iteration

  if (!is.finite(effect_mult) || effect_mult < 0 || effect_mult > 1) {
    stop("Poisson thinning requires effect_mult in [0, 1].")
  }

  # 1) sample IDs
  sampled_treated <- sample(treated_units, size = n_treated, replace = FALSE)
  sampled_untreated <- sample(untreated_units, size = n_untreated, replace = FALSE)

  sampled_ids <- c(sampled_treated, sampled_untreated)

  # 2) subset panel (all years for sampled units), then apply effect only in this sample
  # Poisson thinning: for treated-years, y_new ~ Binomial(y, p = effect_mult)
  sample_df <- sim_with_treat %>%
    filter(study_id %in% sampled_ids) %>%
    mutate(
      yearly_cites_new = if_else(
        D_lag == 1L,
        rbinom(n(), size = pmax(yearly_cites, 0), prob = effect_mult),
        yearly_cites
      )
    )


  # 3) run fect with try()
  fit <- try(
    fect(
      yearly_cites_new ~ D_lag,
      data = sample_df,
      index = c("study_id", "year"),
      method = "ife",
      r = 2,
      se = TRUE,
      parallel = parallel_fect,
      nboots = 50,
      min.T0 = 3
    ),
    silent = TRUE
  )

  if (inherits(fit, "try-error")) {
    return(tibble(att = NA_real_, p.value = NA_real_))
  }

  att <- NA_real_
  pval <- NA_real_
  if (!is.null(fit$est.avg) && "ATT.avg" %in% colnames(fit$est.avg)) {
    att <- as.numeric(fit$est.avg[1, "ATT.avg"])
  } else if (!is.null(fit$att.avg)) {
    att <- as.numeric(fit$att.avg[1])
  }
  if (!is.null(fit$est.avg) && "p.value" %in% colnames(fit$est.avg)) {
    pval <- as.numeric(fit$est.avg[1, "p.value"])
  }

  tibble(att = att, p.value = pval)
}
```

# run the analysis, takes long time

```{r}
effect_sizes <- c(0.75, 0.80, 0.85, 0.90)
treated_sizes <- c(800)
n_iter <- 100

set.seed(225)

run_power_sims <- function(parallel_strategy = c("outer", "inner"),
                           trim_pre = 8,
                           trim_post = 8) {

  parallel_strategy <- match.arg(parallel_strategy)

  # trim to avoid too long computation
  sim_trimmed <- sim_with_treat %>%
    mutate(
      event_time = if_else(!is.na(treat_year),
        year - treat_year,
        NA_integer_
      )
    ) %>%
    filter(
      is.na(event_time) |
        (
          (is.null(trim_pre) | event_time >= -trim_pre) &
            (is.null(trim_post) | event_time <= trim_post)
        )
    )


  ncores <- max(1L, as.integer(future::availableCores()) - 2L)
  old_plan <- future::plan()
  on.exit(future::plan(old_plan), add = TRUE)
  future::plan(future::multisession, workers = ncores)

  sim_grid <- tidyr::expand_grid(
    effect_mult = effect_sizes,
    n_treated   = treated_sizes,
    iter        = 1:n_iter
  ) %>%
    dplyr::mutate(n_untreated = 10L * n_treated)

  if (parallel_strategy[1] == "outer") {
    progressr::with_progress({
      # map_dfr(
      furrr::future_map_dfr(
        1:nrow(sim_grid),
        function(i) {
          row <- sim_grid[i, ]
          out <- run_one_fect_sim(
            iter            = row$iter,
            sim_with_treat  = sim_trimmed,
            treated_units   = treated_units,
            untreated_units = untreated_units,
            n_treated       = row$n_treated,
            n_untreated     = row$n_untreated,
            effect_mult     = row$effect_mult,
            parallel_fect   = FALSE
          )

          dplyr::mutate(
            out,
            effect_mult = row$effect_mult,
            iter        = row$iter,
            n_treated   = row$n_treated,
            n_untreated = row$n_untreated
          )
        },
        .options = furrr::furrr_options(seed = TRUE),
        .progress = TRUE
      )
    })
  } else {
    progressr::with_progress({
      p <- progressr::progressor(along = 1:nrow(sim_grid))

      map_dfr(
        1:nrow(sim_grid),
        function(i) {
          p(sprintf("FECT fit %d/%d", i, nrow(sim_grid)))
          row <- sim_grid[i, ]
          out <- run_one_fect_sim(
            iter            = row$iter,
            sim_with_treat  = sim_trimmed,
            treated_units   = treated_units,
            untreated_units = untreated_units,
            n_treated       = row$n_treated,
            n_untreated     = row$n_untreated,
            effect_mult     = row$effect_mult,
            parallel_fect   = TRUE
          )

          dplyr::mutate(
            out,
            effect_mult = row$effect_mult,
            iter        = row$iter,
            n_treated   = row$n_treated,
            n_untreated = row$n_untreated
          )
        },
        .progress = TRUE
      )
    })
  }
}

# this is closer to the intended use of progressr - outside functions. That way it works within the Rmd files and also when knitting
progressr::handlers(handler_rstudio())
sim_result <- run_power_sims(parallel_strategy = "inner")

power_summary <- sim_result %>%
  dplyr::group_by(effect_mult, n_treated, n_untreated) %>%
  dplyr::summarise(
    power = mean(p.value <= 0.05 & att < 0, na.rm = TRUE),
    n_success = sum(!is.na(p.value)),
    avg_att = mean(att, na.rm = TRUE),
    .groups = "drop"
  )

power_summary

# write_csv(sim_result,"power_sim_pilot.csv")

# sim_res <- sim_results_all
```
