---
title: "_0_power"
author: "Josefina Weinerova"
date: "2025-11-21"
output: html_document
---
```{r}
# install.packages("openalexR")
if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  openalexR, cowplot, tidyverse, httr, jsonlite,
  citecorp, rcrossref, purrr, dplyr, tibble, fect, furrr, future, data.table,
  panelView, rscopus, httr2, plotly, MASS, GGally, metafor, progressr, readxl, openxlsx
)

# loading data
analysis_df <- readxl::read_excel("data/analysis_df.xlsx")
cos_sample <- readxl::read_excel("2025-02-07_COSdata_validated_delivered.xlsx")
all_matched_articles <- readxl::read_excel("data/all_matched_articles_pilot.xlsx")


plan(multisession, workers = availableCores() - 1) # @Josefina: To enable parallel processing. If you are on Mac, you can also try multicore instead of multisession - faster but less stable

my_seed <- 123
```


## Power
## Data-Generating Process (Simulation Outline)

1. Get summary metrics from a sample of control studies
- first year citations
- cumulative last year citations
- number of years
- average growth (average number of citations per year)
- early growth (average number of citations in first 3 years)

2. Create Gaussian multivariate distribution by first ranking the individual metrics and then creating normal distribution
- calculate correlation among different summary metrics
- in the normal space simulate 100,000 new studies by sampling from the distribution while keeping the correlation intact
- transport back to original space by inverse function

# @Josefina: clarify above that you did not simulate collinear variables (e.g., total citations and average growth are collinear)

3. Use the 5 summary metrics to create citation trajectories of the simulated studies

4. select 10,000 studies as treated studies
- select year of treatment start; effect starts 1 year later (use `D_lag`)
- apply effect via *Poisson thinning*: for treated-years, `y_new ~ Binomial(y, p = effect_mult)`, which preserves the discrete observed trajectory while reducing expected counts

5. run fect analysis 100 times for each effect size, collect p.values and ATTs

# what do real citation trajectories look like   

```{r}
analysis_df$year <- as.numeric(analysis_df$year)
treated_dois <- cos_sample$doi_o
analysis_df <- analysis_df %>%
  dplyr::group_by(doi_queried) %>%
  complete(year = full_seq(c(min(year), max(year)), 1), fill = list(n_citations = 0)) %>%
  ungroup()

doi_sample <- analysis_df %>%
  dplyr::distinct(doi_queried) %>%
  dplyr::filter(!(doi_queried %in% treated_dois)) %>%
  dplyr::pull(doi_queried)

doi_sample <- sample(doi_sample, size = min(10000, length(doi_sample)))

# 1) Compute cumulative citations per doi over time for a sample of 200 dois
analysis_df_cum <- analysis_df %>% filter(doi_queried %in% doi_sample)
range(analysis_df_cum$year)
# checking for weird start years
# Josefina: commented out as analysis_df_cum already contains publication year
# Lukas: This is checking for years that are impossibly early. analysis_df_cum has year but not publication year - this removes years that are impossibly early
analysis_df_cum <- analysis_df_cum %>%
  left_join(
    all_matched_articles %>%
      dplyr::select(doi, publication_year),
    by = c("doi_queried" = "doi")
  )

analysis_df_cum <- analysis_df_cum %>% filter(year >= publication_year)
analysis_df_cum <- analysis_df_cum %>%
  arrange(doi_queried, year) %>%
  dplyr::group_by(doi_queried) %>%
  mutate(cum_citations = cumsum(dplyr::coalesce(n_citations, 0))) %>%
  ungroup()

analysis_df_cum$year <- as.numeric(analysis_df_cum$year)
analysis_df_cum <- analysis_df_cum %>%
  dplyr::group_by(doi_queried) %>%
  mutate(time_since_pub = year - publication_year + 1) %>%
  ungroup()
summary(analysis_df_cum$cum_citations)
# not normalized
ggplot(
  analysis_df_cum,
  aes(
    x = time_since_pub,
    y = cum_citations,
    group = doi_queried
  )
) +
  geom_line(alpha = 0.1) +
  labs(
    x = "Year",
    y = "Cumulative citations",
    color = "DOI",
    title = "Cumulative citation growth over time"
  ) +
  theme_minimal() +
  scale_y_log10()


# @Josefina: these numbers seem too high - median cumuulative citations are 54 at the end of the period, while I have usually seen *much* lower numbers, e.g.,  https://lucbeaulieu.com/2015/11/19/how-many-citations-are-actually-a-lot-of-citations/. This could of course be right as replication targets are not selected randomly - but can you please double-check and then explicitly note that divergence? [And consider whetre to place that in the article?] Also, is our method excluding articles with 0 citations? We have 12 of them, which seems really low ...
```
#plotting distribution of parameters

```{r}
citation_summary <- analysis_df_cum %>%
  dplyr::group_by(doi_queried) %>%
  arrange(time_since_pub) %>%
  mutate(
    yearly_cites = cum_citations - lag(cum_citations, default = 0)
  ) %>%
  dplyr::summarise(
    first_year_citations = yearly_cites[time_since_pub == 1],
    max_cumulative = max(cum_citations),
    years_observed = max(time_since_pub),
    avg_yearly_growth = mean(yearly_cites),
    early_growth = mean(yearly_cites[time_since_pub <= min(3, years_observed)]),
    linear_slope = coef(lm(cum_citations ~ time_since_pub))[2],
    log_slope = coef(lm(log1p(cum_citations) ~ time_since_pub))[2]
  )

citation_summary <- na.omit(citation_summary)
plot_ly(
  citation_summary,
  x = ~first_year_citations,
  y = ~max_cumulative,
  z = ~years_observed,
  color = ~doi_queried, # color by DOI
  colors = "Dark2",
  type = "scatter3d",
  mode = "markers"
)

citation_summary <- as.data.frame(citation_summary)

summary(citation_summary)
cor(citation_summary %>% dplyr::select(-doi_queried))


ggpairs(citation_summary, columns = 2:8)
```


#sampling from real distributions
```{r}
vars <- citation_summary %>%
  dplyr::select(
    first_year_citations,
    max_cumulative,
    years_observed,
    avg_yearly_growth,
    early_growth
  )

n <- nrow(vars)

# empirical CDF via ranks, then map to normal
to_normal <- function(x) {
  u <- (rank(x, ties.method = "average") - 0.5) / n # in (0,1)
  qnorm(u) # ≈ N(0,1)
}

Z <- vars %>%
  mutate(across(everything(), to_normal)) %>%
  as.matrix()


Sigma <- cor(Z)

n_sim <- 100000

Z_sim <- mvrnorm(n = n_sim, mu = rep(0, ncol(Z)), Sigma = Sigma)
colnames(Z_sim) <- colnames(vars)


# helper: empirical quantile function
inv_ecdf <- function(x, u) {
  # u in (0,1); use type=8 interpolation
  quantile(x, probs = u, type = 8, na.rm = TRUE)
}

U_sim <- pnorm(Z_sim) # now values in (0,1)

sim_vars <- matrix(NA, nrow = n_sim, ncol = ncol(vars))
colnames(sim_vars) <- colnames(vars)

for (j in seq_along(vars)) {
  sim_vars[, j] <- inv_ecdf(vars[[j]], U_sim[, j])
}

sim_summaries <- as.data.frame(sim_vars)

summary(sim_summaries)
```


#simulate from summaries
```{r}
simulate_yearly_from_summary <- function(
  Time,
  total,
  first1,
  early_growth = NULL,
  last_year = 2025
) {
  # Edge cases:
  Time <- max(1, round(Time))
  total <- max(0, round(total))
  first1 <- round(first1)

  # If no citations at all
  if (total == 0) {
    years <- seq(last_year - Time + 1, last_year)
    return(tibble(
      year = years,
      yearly_cites = rep(0, Time),
      cum_cites = rep(0, Time)
    ))
  }

  # If only 1 year observed, everything happens in that year
  if (Time == 1) {
    years <- last_year
    return(tibble(
      year = years,
      yearly_cites = total,
      cum_cites = total
    ))
  }

  # Ensure first-year citations are between 0 and total
  first1 <- max(0, min(first1, total))

  # ---- Baseline linear shape ----
  # yearly(t) = a + b*t
  b <- 2 * (total - Time * first1) / (Time * (Time - 1))
  a <- first1 - b

  t_seq <- 1:Time
  lam <- a + b * t_seq

  # ---- Shape adjustment using ONLY early_growth ----
  # derive the implied average yearly citations from (total, T)
  avg_growth_derived <- total / Time

  if (!is.null(early_growth) && is.finite(early_growth) && avg_growth_derived > 0) {
    # early_ratio > 1 -> front-loaded, < 1 -> back-loaded
    early_ratio <- early_growth / avg_growth_derived

    # clamp to avoid crazy numbers
    early_ratio <- pmin(pmax(early_ratio, 0.3), 3)

    # smooth warping parameter
    k <- log(early_ratio) / max(Time - 1, 1)

    # weights over time
    w <- exp(-k * (t_seq - 1))

    lam <- lam * w
  }

  # Enforce non-negativity
  lam[lam < 0] <- 0

  # If everything went to zero, fall back to uniform
  if (sum(lam) == 0) {
    lam <- rep(total / Time, Time)
  } else {
    # Rescale lambdas so they sum exactly to total in expectation
    lam <- lam * (total / sum(lam))
  }

  # Generate integer yearly citations (Poisson noise)
  yearly <- rpois(Time, lambda = lam)

  # Avoid the degenerate all-zero case
  if (sum(yearly) == 0) {
    yearly <- round(lam)
  }

  # Adjust mismatch so sum(yearly) == total
  diff <- total - sum(yearly)
  if (diff != 0) {
    idx <- sample(1:Time, size = abs(diff), replace = TRUE)
    for (i in idx) {
      yearly[i] <- yearly[i] + sign(diff)
    }
    yearly[yearly < 0] <- 0
  }

  cum <- cumsum(yearly)
  years <- seq(last_year - Time + 1, last_year)

  tibble(
    year = years,
    yearly_cites = yearly,
    cum_cites = cum
  )
}
```

#apply to simulated studies
```{r}
sim_summaries <- sim_summaries %>%
  mutate(doi_queried = row_number())


handlers(handler_rstudio())

with_progress({
  simulated_trajectories <- sim_summaries %>%
    dplyr::select(
      doi_queried,
      years_observed,
      max_cumulative,
      first_year_citations,
      early_growth
    ) %>%
    future_pmap_dfr(
      function(doi_queried,
               years_observed,
               max_cumulative,
               first_year_citations,
               early_growth) {
        simulate_yearly_from_summary(
          Time = years_observed,
          total = max_cumulative,
          first1 = first_year_citations,
          early_growth = early_growth,
          last_year = 2025
        ) %>%
          mutate(study_id = doi_queried, .before = 1)
      },
      .progress = TRUE,
      .options = furrr_options(seed = my_seed)
    )
})

range(simulated_trajectories$year)
simulated_trajectories <- simulated_trajectories %>%
  dplyr::group_by(study_id) %>%
  mutate(
    start_year = min(year),
    time_since_pub = year - start_year + 1
  ) %>%
  ungroup()

# not normalized
ggplot(
  simulated_trajectories,
  aes(
    x = time_since_pub,
    y = cum_cites,
    group = study_id
  )
) +
  geom_line(alpha = 0.1) +
  labs(
    x = "Year",
    y = "Cumulative citations",
    color = "DOI",
    title = "Simulated cumulative citation growth over time"
  ) +
  theme_minimal() +
  scale_y_log10()

# @Josefina: cum_cites here is substantially higher than in analysis_df_cum - median is 73 rather than 54. Is that expected?
```

```{r}
# min_post_years <- 5L  # require at least 5 years with D==1 @Josefina: why? Don't we only have *pre* restrictions? - @This was one of the things I did when the power analysis stoped

study_info <- simulated_trajectories %>%
  dplyr::group_by(study_id) %>%
  dplyr::summarise(
    start_year = min(year),
    end_year = max(year),
    .groups = "drop"
  )


study_info <- study_info %>%
  mutate(
    treat_from = pmax(start_year + 5, 2010L),
    treat_to   = 2024L
  ) %>%
  filter(treat_from <= treat_to)


summary(study_info)

# sample treated studies
set.seed(123)

treated_ids <- sample(study_info$study_id, size = 10000)

# assign treatment
treat_info <- study_info %>%
  filter(study_id %in% treated_ids) %>%
  rowwise() %>%
  mutate(
    treat_year = if_else(
      treat_from == treat_to,
      treat_from,
      sample(treat_from:treat_to, size = 1)
    )
  ) %>%
  ungroup() %>%
  dplyr::select(study_id, treat_year)


sim_with_treat <- simulated_trajectories %>%
  left_join(treat_info, by = "study_id") %>%
  mutate(
    treated = !is.na(treat_year),
    # D = 1 in the treatment year and all years after
    D = if_else(
      is.na(treat_year), # untreated studies
      0L,
      if_else(year >= treat_year, 1L, 0L)
    )
  )

sim_with_treat <- sim_with_treat %>%
  dplyr::group_by(study_id) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    D_lag = dplyr::lag(D, default = 0L) # D_{t-1}
  ) %>%
  ungroup()

## commented for the full run, this part is already implemented below, here it was used as checks for the effect implementation before the run

## Previous approach (stochastic rounding) kept for reference; main simulation now uses Poisson thinning
# sim_with_treat <- sim_with_treat %>%
#   mutate(
#     lambda_baseline = pmax(yearly_cites, 0),
#     lambda_treated  = if_else(
#       D_lag == 1L,                 # or D == 1L if you want instantaneous effect
#       0.8 * lambda_baseline,
#       lambda_baseline
#     ),
#     lambda_treated = pmax(lambda_treated, 0),   # safety
#
#     # stochastic rounding:
#     floor_val = floor(lambda_treated),
#     frac      = lambda_treated - floor_val,     # e.g. 0.80 for 3.80
#     frac      = pmin(pmax(frac, 0), 1),         # clamp to [0,1]
#
#     # add Bernoulli(frac): 1 with prob = frac, 0 otherwise
#     yearly_cites_new = floor_val + rbinom(n(), size = 1, prob = frac)
#   )
```
#plot after adding treatment
```{r}
set.seed(123)
example_effect_mult <- 0.9
sim_with_treat_example <- sim_with_treat %>%
  mutate(
    yearly_cites_new = if_else(
      D_lag == 1L,
      rbinom(n(), size = pmax(yearly_cites, 0), prob = example_effect_mult), 
      yearly_cites 
    )
  ) %>%
  dplyr::group_by(study_id) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(cum_cites_new = cumsum(yearly_cites_new)) %>%
  ungroup()

# did not have publication year and thus time_since_pub in the data - check that my code below is correct
sim_with_treat_example <- sim_with_treat_example %>%
  group_by(study_id) %>%
  mutate(
    publication_year = min(year),
    time_since_pub = year - publication_year + 1
  )

ggplot(
  sim_with_treat_example,
  aes(
    x = time_since_pub,
    y = cum_cites_new,
    group = study_id, colour = factor(D_lag)
  )
) +
  geom_line(alpha = 0.2) +
  labs(
    x = "Year",
    y = "Cumulative citations",
    color = "Treatment status", # @Josefina: renamed from DOI - right?
    title = "Simulated cumulative citation growth over time"
  ) +
  theme_minimal() +
  scale_y_log10()

a <- sim_with_treat_example %>%
  dplyr::group_by(study_id) %>%
  dplyr::filter(any(D == 1L)) %>%
  dplyr::filter(study_id == 5)

ggplot(a, aes(x = year, group = study_id)) +
  geom_line(aes(y = yearly_cites, colour = "baseline")) +
  geom_line(aes(y = yearly_cites_new, colour = "thinned")) +
  labs(y = "Yearly citations", colour = NULL)

postlag_by_id <- sim_with_treat %>%
  filter(treated) %>%
  group_by(study_id) %>%
  summarise(
    end_year = max(year),
    treat_year = first(treat_year),
    postlag_years = sum(D_lag == 1L),
    .groups = "drop"
  )

# @Josefina: this is likely longer than in FReD - but not sure if it matters
summary(postlag_by_id$postlag_years)
```


```{r}
unit_treat_status <- sim_with_treat %>%
  dplyr::group_by(study_id) %>%
  dplyr::summarise(
    ever_treated = any(D == 1L),
    .groups = "drop"
  )

treated_units <- unit_treat_status$study_id[unit_treat_status$ever_treated]
untreated_units <- unit_treat_status$study_id[!unit_treat_status$ever_treated]
```

```{r}
run_one_fect_sim <- function(iter,
                             sim_with_treat,
                             treated_units,
                             untreated_units,
                             n_treated = 1000,
                             n_untreated = 10000,
                             effect_mult = 0.9,
                             parallel_fect = TRUE) {
  set.seed(iter) # optional: reproducible per iteration

  if (!is.finite(effect_mult) || effect_mult < 0 || effect_mult > 1) {
    stop("Poisson thinning requires effect_mult in [0, 1].")
  }

  # 1) sample IDs
  sampled_treated <- sample(treated_units, size = n_treated, replace = FALSE)
  sampled_untreated <- sample(untreated_units, size = n_untreated, replace = FALSE)

  sampled_ids <- c(sampled_treated, sampled_untreated)

  # 2) subset panel (all years for sampled units), then apply effect only in this sample
  # Poisson thinning: for treated-years, y_new ~ Binomial(y, p = effect_mult)
  sample_df <- sim_with_treat %>%
    filter(study_id %in% sampled_ids) %>%
    mutate(
      yearly_cites_new = if_else(
        D_lag == 1L,
        rbinom(n(), size = pmax(yearly_cites, 0), prob = effect_mult),
        yearly_cites
      )
    )


  # 3) run fect with try()
  fit <- try(
    fect(
      yearly_cites_new ~ D_lag,
      data = sample_df,
      index = c("study_id", "year"),
      method = "ife",
      r = 2,
      se = TRUE,
      parallel = parallel_fect,
      nboots = 50,
      min.T0 = 3
    ),
    silent = TRUE
  )

  if (inherits(fit, "try-error")) {
    return(tibble(att = NA_real_, p.value = NA_real_))
  }

  att <- NA_real_
  pval <- NA_real_
  if (!is.null(fit$est.avg) && "ATT.avg" %in% colnames(fit$est.avg)) {
    att <- as.numeric(fit$est.avg[1, "ATT.avg"])
  } else if (!is.null(fit$att.avg)) {
    att <- as.numeric(fit$att.avg[1]) # @Josefina: what is this conditional for? fect output should be consistent?
  }
  if (!is.null(fit$est.avg) && "p.value" %in% colnames(fit$est.avg)) {
    pval <- as.numeric(fit$est.avg[1, "p.value"])
  }

  tibble(att = att, p.value = pval)
}
```

# run the analysis, takes long time

```{r}
effect_sizes <- c(0.75, 0.80, 0.85, 0.90)
treated_sizes <- c(800)
n_iter <- 100

set.seed(225)

run_power_sims <- function(parallel_strategy = c("outer", "inner"),
                           trim_pre = 8,
                           trim_post = 8) {
  # would expect inner to be better as fect is more computationally intensive

  parallel_strategy <- match.arg(parallel_strategy)

  # trim to avoid too long computation
  sim_trimmed <- sim_with_treat %>%
    mutate(
      event_time = if_else(!is.na(treat_year),
        year - treat_year,
        NA_integer_
      )
    ) %>%
    filter(
      is.na(event_time) |
        (
          (is.null(trim_pre) | event_time >= -trim_pre) &
            (is.null(trim_post) | event_time <= trim_post)
        )
    )


  ncores <- max(1L, as.integer(future::availableCores()) - 2L)
  old_plan <- future::plan()
  on.exit(future::plan(old_plan), add = TRUE)
  future::plan(future::multisession, workers = ncores)

  sim_grid <- tidyr::expand_grid(
    effect_mult = effect_sizes,
    n_treated   = treated_sizes,
    iter        = 1:n_iter
  ) %>%
    dplyr::mutate(n_untreated = 10L * n_treated)

  if (parallel_strategy[1] == "outer") {
    progressr::with_progress({
      # map_dfr(
      furrr::future_map_dfr(
        1:nrow(sim_grid),
        function(i) {
          row <- sim_grid[i, ]
          out <- run_one_fect_sim(
            iter            = row$iter,
            sim_with_treat  = sim_trimmed,
            treated_units   = treated_units,
            untreated_units = untreated_units,
            n_treated       = row$n_treated,
            n_untreated     = row$n_untreated,
            effect_mult     = row$effect_mult,
            parallel_fect   = FALSE
          )

          dplyr::mutate(
            out,
            effect_mult = row$effect_mult,
            iter        = row$iter,
            n_treated   = row$n_treated,
            n_untreated = row$n_untreated
          )
        },
        .options = furrr::furrr_options(seed = TRUE),
        .progress = TRUE
      )
    })
  } else {
    progressr::with_progress({
      p <- progressr::progressor(along = 1:nrow(sim_grid))

      map_dfr(
        1:nrow(sim_grid),
        function(i) {
          p(sprintf("FECT fit %d/%d", i, nrow(sim_grid)))
          row <- sim_grid[i, ]
          out <- run_one_fect_sim(
            iter            = row$iter,
            sim_with_treat  = sim_trimmed,
            treated_units   = treated_units,
            untreated_units = untreated_units,
            n_treated       = row$n_treated,
            n_untreated     = row$n_untreated,
            effect_mult     = row$effect_mult,
            parallel_fect   = TRUE
          )

          dplyr::mutate(
            out,
            effect_mult = row$effect_mult,
            iter        = row$iter,
            n_treated   = row$n_treated,
            n_untreated = row$n_untreated
          )
        },
        .progress = TRUE
      )
    })
  }
}

# this is closer to the intended use of progressr - outside functions. That way it works within the Rmd files and also when knitting
progressr::handlers(handler_rstudio())
sim_result <- run_power_sims(parallel_strategy = "inner")

#I believe we agreed to reduce min.T0 to r+1? Can you check and change that?

power_summary <- sim_result %>%
  dplyr::group_by(effect_mult, n_treated, n_untreated) %>%
  dplyr::summarise(
    power = mean(p.value <= 0.05 & att < 0, na.rm = TRUE),
    n_success = sum(!is.na(p.value)),
    avg_att = mean(att, na.rm = TRUE),
    .groups = "drop"
  )

power_summary

# write_csv(sim_result,"power_sim_pilot.csv")

# sim_res <- sim_results_all
```

# power for metafor - meta-regression

```{r}
# set.seed(125) # so it runs once reproducibly
# 
# # ---- global parameters ----
# I <- 500
# mu <- -5
# tau2 <- 20
# 
# beta_m1 <- 1
# beta_m2 <- 0
# beta_m3_levels <- c(0, 0.5, 1.8, 1.0, 1.2, 1.5, -0.5, -0.8, -1.0, -2)
# 
# beta_c1 <- -0.5
# beta_c2 <- 0
# beta_c3 <- 0.6
# 
# # ---- simulate moderators once ----
# mods_study <- data.frame(
#   study = 1:I,
#   m1    = rbinom(I, 1, 0.5),
#   m2    = rbinom(I, 1, 0.4),
#   m3    = sample(1:10, I, replace = TRUE),
#   c1    = rnorm(I, mean = 0, sd = 1),
#   c2    = rnorm(I, mean = 10, sd = 2),
#   c3    = rnorm(I, mean = 50, sd = 10)
# )
# 
# # treat m3 as a factor so metafor creates dummies vs level 1
# mods_study$m3 <- factor(mods_study$m3, levels = 1:10)
# 
# # ---- simulate sampling variances (vi) once ----
# # choose something plausible; you can replace this with your own mechanism
# mods_study$vi <- runif(I, min = 0.5, max = 5)
# 
# # ---- generate true effects theta_i ----
# m3_effect <- beta_m3_levels[as.integer(mods_study$m3)]
# 
# theta <- mu +
#   beta_m1 * mods_study$m1 +
#   beta_m2 * mods_study$m2 +
#   m3_effect +
#   beta_c1 * mods_study$c1 +
#   beta_c2 * mods_study$c2 +
#   beta_c3 * mods_study$c3 +
#   rnorm(I, mean = 0, sd = sqrt(tau2)) # between-study heterogeneity
# se_fun <- function(n) {
#   x <- rlnorm(n, meanlog = log(3), sdlog = 0.6)
#   x[x > 16] <- 16
#   x
# }
# 
# mods_study$sei <- se_fun(I)
# mods_study$vi <- mods_study$sei^2
# # ---- observed study effects yi ----
# mods_study$yi <- theta + rnorm(I, mean = 0, sd = sqrt(mods_study$vi))
# 
# # ---- fit ONE metafor model ----
# fit <- rma(yi, vi,
#   mods = ~ m1 + m2 + m3 + c1 + c2 + c3,
#   data = mods_study,
#   method = "REML"
# )
# 
# summary(fit)
# 
# 
# fit$pval[2]
```
# linear metafor model
```{r}
# set.seed(125)
# 
# # ---- global parameters ----
# I <- 500 # number of studies
# mu <- -5 # true average ATT
# tau2 <- 20 # between-study variance
# 
# beta_m2 <- 0
# beta_m3_levels <- c(0, 0.5, 1.8, 1.0, 1.2, 1.5, -0.5, -0.8, -1.0, -2)
# 
# beta_c1 <- -0.5
# beta_c2 <- 0
# beta_c3 <- 0.6
# 
# # ---- sampling SE function ----
# se_fun <- function(n) {
#   x <- rlnorm(n, meanlog = log(3), sdlog = 0.6)
#   x[x > 16] <- 16
#   x
# }
# 
# # ---- simulation settings ----
# m1_grid <- c(0.5, 1, 1.5, 2)
# n_sims <- 400
# 
# # (If you prefer balanced 25 each, use instead:)
# m1_by_run <- sample(rep(m1_grid, each = n_sims / length(m1_grid)))
# 
# # ---- results container ----
# results <- data.frame(
#   iter        = seq_len(n_sims), # counter
#   beta_m1     = m1_by_run, # true m1 effect used
#   pval_m1     = NA_real_ # fit$pval[2]
# )
# 
# for (i in seq_len(n_sims)) {
#   beta_m1 <- m1_by_run[i]
# 
#   # simulate moderators
#   mods_study <- data.frame(
#     study = 1:I,
#     m1    = rbinom(I, 1, 0.5),
#     m2    = rbinom(I, 1, 0.4),
#     m3    = sample(1:10, I, replace = TRUE),
#     c1    = rnorm(I, mean = 0, sd = 1),
#     c2    = rnorm(I, mean = 10, sd = 2),
#     c3    = rnorm(I, mean = 50, sd = 10)
#   )
#   mods_study$m3 <- factor(mods_study$m3, levels = 1:10)
# 
#   # sampling variances
#   mods_study$sei <- se_fun(I)
#   mods_study$vi <- mods_study$sei^2
# 
#   # true study effects
#   m3_effect <- beta_m3_levels[as.integer(mods_study$m3)]
# 
#   theta <- mu +
#     beta_m1 * mods_study$m1 +
#     beta_m2 * mods_study$m2 +
#     m3_effect +
#     beta_c1 * mods_study$c1 +
#     beta_c2 * mods_study$c2 +
#     beta_c3 * mods_study$c3 +
#     rnorm(I, mean = 0, sd = sqrt(tau2))
# 
#   # observed effects
#   mods_study$yi <- theta + rnorm(I, mean = 0, sd = sqrt(mods_study$vi))
# 
#   # fit meta-regression and store p-value for m1
#   fit <- tryCatch(
#     rma(yi, vi,
#       mods = ~ m1 + m2 + m3 + c1 + c2 + c3,
#       data = mods_study,
#       method = "REML"
#     ),
#     error = function(e) NULL
#   )
# 
#   if (!is.null(fit)) {
#     results$pval_m1[i] <- fit$pval[2]
#   }
# 
#   # optional progress print
#   message("Completed ", i, "/", n_sims)
# }
# 
# results
# 
# alpha <- 0.05
# 
# power_mf <- results %>%
#   dplyr::group_by(beta_m1) %>%
#   summarise(
#     n = sum(!is.na(pval_m1)),
#     pct_below_0.05 = 100 * mean(pval_m1 < alpha, na.rm = TRUE)
#   ) %>%
#   arrange(beta_m1)
# 
# power_mf
```
# mixed effects metafor model
```{r}
# model types to modify runs
# m1_only   = ~ m1,
# m1_m2_c1  = ~ m1 + m2 + c1
# full      = ~ m1 + m2 + m3 + c1 + c2 + c3

set.seed(125)

# ---- global parameters ----
I <- 1000
mu <- -2
tau2 <- 20

beta_m2 <- 0.2 # same journal
beta_m3_levels <- c(0, 0.5, 1.8, 1.0, 1.2, 1.5, -0.5, -0.8, -1.0, -2) #subject field
beta_m4_levels <- c(0,-0.8,0.3) #single publication vs many labs vs unpublished
beta_c1 <- -0.5 # time between original and replic
beta_c2 <- 1 #h-index
beta_c3 <- 0.6 #snip

se_fun <- function(n) {
  x <- rlnorm(n, meanlog = log(3), sdlog = 0.6)
  x[x > 16] <- 16
  x
}

# ---- simulation settings ----
m1_grid <- c(0.5, 1, 1.5, 2)
n_sims <- 400
m1_by_run <- sample(rep(m1_grid, each = n_sims / length(m1_grid)))

results <- data.frame(
  iter    = seq_len(n_sims),
  beta_m1 = m1_by_run,
  pval_m1 = NA_real_
)

J <- 4 # effects per study

for (i in seq_len(n_sims)) {
  beta_m1 <- m1_by_run[i]

  # ---- study-level moderators ----
  mods_study <- data.frame(
    study = factor(seq_len(I)),
    m1    = rbinom(I, 1, 0.5),
    m2    = rbinom(I, 1, 0.4),
    m3    = factor(sample(1:10, I, replace = TRUE), levels = 1:10),
    c1    = rnorm(I, 0, 1),
    c2    = rnorm(I, 10, 2),
    c3    = rnorm(I, 50, 10)
  )

  # ---- expand to effect-level data (4 rows per study) ----
  sim <- mods_study[rep(seq_len(I), each = J), ]
  sim$effect <- factor(rep(seq_len(J), times = I))

  k <- nrow(sim)

  # sampling variances at effect level
  sim$sei <- se_fun(k)
  sim$vi <- sim$sei^2

  # study random effects (shared across the 4 effects)
  delta_i <- rnorm(I, 0, sqrt(tau2))
  sim$delta_i <- rep(delta_i, each = J)

  # m3 fixed effect contribution
  sim$m3_effect <- beta_m3_levels[as.integer(sim$m3)]

  # linear predictor excluding m1 (so we can vary beta_m1 easily)
  base_lp <- mu +
    sim$delta_i +
    beta_m2 * sim$m2 +
    sim$m3_effect +
    beta_c1 * sim$c1 +
    beta_c2 * sim$c2 +
    beta_c3 * sim$c3

  # observed effect sizes
  sim$yi <- base_lp + beta_m1 * sim$m1 + rnorm(k, 0, sim$sei)

  # ---- fit rma.mv with random intercept for study ----
  fit <- tryCatch(
    rma.mv(
      yi = yi, V = vi,
      mods = ~ m1 + m2 + m3 + c1 + c2 + c3,
      random = ~ 1 | study,
      data = sim,
      method = "REML",
      control = list(mfmaxit = 200, sigma2.init = tau2)
    ),
    error = function(e) NULL
  )

  if (!is.null(fit)) {
    pvec <- fit$pval
    names(pvec) <- rownames(fit$b)
    results$pval_m1[i] <- unname(pvec["m1"])
  }

  message("Completed ", i, "/", n_sims)
}

results

alpha <- 0.05

power_mva <- results %>%
  dplyr::group_by(beta_m1) %>%
  summarise(
    n = sum(!is.na(pval_m1)),
    pct_below_0.05 = 100 * mean(pval_m1 < alpha, na.rm = TRUE)
  ) %>%
  arrange(beta_m1)

power_mva
```
```{r}
# model types to modify runs
# m1_only   = ~ m1,
# m1_m2_c1  = ~ m1 + m2 + c1
# full      = ~ m1 + m2 + m3 + m4 + c1 + c2 + c3 + yrs_rep + yrs_rep2

set.seed(125)

# ---- global parameters ----
I <- 1000
mu <- -2
tau2 <- 20

beta_m2 <- 0.2 # same journal
beta_m3_levels <- c(0, 0.5, 1.8, 1.0, 1.2, 1.5, -0.5, -0.8, -1.0, -2) #subject field
beta_m4_levels <- c(0, -0.8, 0.3) #single publication vs many labs vs unpublished

beta_c1 <- -0.5 # time between original and replic
beta_c2 <- 1    # h-index
beta_c3 <- 0.6  # snip

# NEW: effect-level "years since replication" moderators (data-generating effects)
beta_yrs_rep  <- -0.15  # +0.15 per year (requested)
beta_yrs_rep2 <- -0.01  # set to whatever you want for the quadratic; keep 0 if you only want linear

se_fun <- function(n) {
  x <- rlnorm(n, meanlog = log(3), sdlog = 0.6)
  x[x > 16] <- 16
  x
}

# ---- simulation settings ----
m1_grid <- c(0.5, 1, 1.5, 2)
n_sims <- 2
m1_by_run <- sample(m1_grid, size = n_sims, replace = TRUE)

results <- data.frame(
  iter    = seq_len(n_sims),
  beta_m1 = m1_by_run,
  pval_m1 = NA_real_
)

J <- 4 # effects per study

for (i in seq_len(n_sims)) {
  beta_m1 <- m1_by_run[i]

  # ---- study-level moderators ----
  mods_study <- data.frame(
    study = factor(seq_len(I)),
    m1    = rbinom(I, 1, 0.5),
    m2    = rbinom(I, 1, 0.4),
    m3    = factor(sample(1:10, I, replace = TRUE), levels = 1:10),
    m4    = factor(sample(1:3,  I, replace = TRUE), levels = 1:3),   # NEW
    c1    = rnorm(I, 0, 1),
    c2    = rnorm(I, 10, 2),
    c3    = rnorm(I, 50, 10)
  )

  # ---- expand to effect-level data (J rows per study) ----
  sim <- mods_study[rep(seq_len(I), each = J), ]

  # effect order within study (kept as factor if you still want it)
  sim$effect <- factor(rep(seq_len(J), times = I), levels = seq_len(J))

  # NEW: years since replication as numeric 0,1,2,3 within each study
  sim$yrs_rep  <- rep(0:(J - 1), times = I)
  sim$yrs_rep2 <- sim$yrs_rep^2

  k <- nrow(sim)

  # sampling variances at effect level
  sim$sei <- se_fun(k)
  sim$vi <- sim$sei^2

  # study random effects (shared across the J effects)
  delta_i <- rnorm(I, 0, sqrt(tau2))
  sim$delta_i <- rep(delta_i, each = J)

  # m3/m4 fixed effect contributions (data-generating)
  sim$m3_effect <- beta_m3_levels[as.integer(sim$m3)]
  sim$m4_effect <- beta_m4_levels[as.integer(sim$m4)]  # NEW

  # linear predictor excluding m1 (so we can vary beta_m1 easily)
  base_lp <- mu +
    sim$delta_i +
    beta_m2 * sim$m2 +
    sim$m3_effect +
    sim$m4_effect +                         # NEW
    beta_c1 * sim$c1 +
    beta_c2 * sim$c2 +
    beta_c3 * sim$c3 +
    beta_yrs_rep  * sim$yrs_rep +           # NEW
    beta_yrs_rep2 * sim$yrs_rep2            # NEW

  # observed effect sizes
  sim$yi <- base_lp + beta_m1 * sim$m1 + rnorm(k, 0, sim$sei)
  
  print(round(cor(sim[, c("m1","m2","c1","c2","c3","yrs_rep","yrs_rep2")]), 3))

  X <- model.matrix(~ m1 + m2 + m3 + m4 + c1 + c2 + c3 + yrs_rep + yrs_rep2, data = sim)
  print(round(cor(X[, -1]), 3))

  # ---- fit rma.mv with random intercept for study ----
  fit <- tryCatch(
    rma.mv(
      yi = yi, V = vi,
      mods = ~ m1 + m2 + m3 + m4 + c1 + c2 + c3 + yrs_rep + yrs_rep2, # UPDATED
      random = ~ 1 | study,
      data = sim,
      method = "REML",
      control = list(mfmaxit = 200, sigma2.init = tau2)
    ),
    error = function(e) NULL
  )

  if (!is.null(fit)) {
    pvec <- fit$pval
    names(pvec) <- rownames(fit$b)
    results$pval_m1[i] <- unname(pvec["m1"])
  }

  message("Completed ", i, "/", n_sims)
}

results

alpha <- 0.05

power_mva <- results %>%
  dplyr::group_by(beta_m1) %>%
  dplyr::summarise(
    n = sum(!is.na(pval_m1)),
    pct_below_0.05 = 100 * mean(pval_m1 < alpha, na.rm = TRUE)
  ) %>%
  dplyr::arrange(beta_m1)

power_mva



```

#parallel script
```{r}
### --- GLOBAL PARAMETERS  --------------------
#
# I   <- 1000        # number of studies
# mu  <- -2         # true average ATT
# tau2 <- 20        # between-study variance
#
# # Other true moderator effects (fixed; m1 will vary)
# beta_m2 <-  0                # binary m2 (null)
# beta_m3_levels <- c(0,       # m3 level 1 (reference)
#                     0.5, 1.8, 1.0, 1.2, 1.5,
#                    -0.5,-0.8,-1.0,-2)  # levels 2–10
#
# beta_c1 <- -0.5              # continuous c1
# beta_c2 <-  0                # continuous c2 (null)
# beta_c3 <-  0.6              # continuous c3
#
# ### --- SAMPLING VARIANCES ----------------------------------------------
# se_fun <- function(n) {
#   x <- rlnorm(n, meanlog = log(3), sdlog = 0.6)
#   x[x > 16] <- 16
#   x
# }
#
# n_sim <- 1   # number of simulation runs per scenario
#
# # Grid of effect sizes for m1 we want to try
# beta_m1_grid <- c(1,2)
#
# # Models to compare
# model_list <- list(
#   m1_only   = ~ m1,                            # 1) m1 only
#   m1_m2_c1  = ~ m1 + m2 + c1,                  # 2) m1 + m2 + c1
#   full      = ~ m1 + m2 + m3 + c1 + c2 + c3    # 3) full model
# )
#
# # 3D array to store ALL p-values for m1:
# # dimensions: [beta_m1 value, model, simulation]
# pvals_m1 <- array(
#   NA_real_,
#   dim = c(length(beta_m1_grid), length(model_list), n_sim),
#   dimnames = list(
#     beta_m1 = paste0("b", beta_m1_grid),
#     model   = names(model_list),
#     sim     = as.character(1:n_sim)
#   )
# )
#
# # --- helper: one simulation replicate (returns matrix [beta, model]) ---
# one_sim <- one_sim_fast <- function(s, I, mu, tau2, beta_m1_grid, model_list,
#                          beta_m2, beta_m3_levels, beta_c1, beta_c2, beta_c3, se_fun) {
#
#   effects_per_study <- sample.int(5, I, replace = TRUE)  # 1:5
#
#   study  <- rep.int(seq_len(I), times = effects_per_study)
#   effect <- sequence(effects_per_study)
#   sim <- data.frame(study = study, effect = effect)
#
#   # study-level moderators
#   mods <- data.frame(
#     m1 = rbinom(I, 1, 0.5),
#     m2 = rbinom(I, 1, 0.4),
#     m3 = sample.int(10, I, replace = TRUE),
#     c1 = rnorm(I, 0, 1),
#     c2 = rnorm(I, 10, 2),
#     c3 = rnorm(I, 50, 10)
#   )
#
#   # map to effect rows by indexing
#   sim$m1 <- mods$m1[sim$study]
#   sim$m2 <- mods$m2[sim$study]
#   sim$c1 <- mods$c1[sim$study]
#   sim$c2 <- mods$c2[sim$study]
#   sim$c3 <- mods$c3[sim$study]
#
#   # m3 at study-level, then map
#   m3_fac_study <- factor(mods$m3, levels = 1:10, labels = paste0("L", 1:10))
#   m3_eff_study <- beta_m3_levels[mods$m3]
#   sim$m3        <- m3_fac_study[sim$study]
#   sim$m3_effect <- m3_eff_study[sim$study]
#
#   # random study effects mapped
#   delta_i <- rnorm(I, 0, sqrt(tau2))
#   sim$delta_i <- delta_i[sim$study]
#
#   # sampling variance and eps once
#   k <- nrow(sim)
#   sim$sei <- se_fun(k)
#   sim$vi  <- sim$sei^2
#   eps <- rnorm(k, 0, sim$sei)
#
#   base_lp <- mu +
#     sim$delta_i +
#     beta_m2 * sim$m2 +
#     sim$m3_effect +
#     beta_c1 * sim$c1 +
#     beta_c2 * sim$c2 +
#     beta_c3 * sim$c3
#
#   # precompute design matrices once per replicate (big win with factor m3)
#   X_list <- lapply(model_list, model.matrix, data = sim)
#
#   out <- matrix(NA_real_, nrow = length(beta_m1_grid), ncol = length(model_list),
#                 dimnames = list(paste0("b", beta_m1_grid), names(model_list)))
#
#   for (b in seq_along(beta_m1_grid)) {
#     beta_m1_cur <- beta_m1_grid[b]
#     yi <- base_lp + beta_m1_cur * sim$m1 + eps
#
#     for (m in seq_along(model_list)) {
#       fit <- try(
#         metafor::rma.mv(yi = yi, V = sim$vi, mods = X_list[[m]],intercept = FALSE,
#                         random = ~ 1 | study, data = sim),
#         silent = TRUE
#       )
#       if (inherits(fit, "try-error")) next
#
#       nm <- rownames(fit$b)
#       p  <- fit$pval
#       names(p) <- nm
#       if ("m1" %in% names(p)) out[b, m] <- p["m1"]
#     }
#   }
#
#   out
# }
#
#
# # --- parallel runner ---
# run_power_parallel <- function(n_sim, workers = NULL) {
#   if (is.null(workers)) workers <- max(1L, future::availableCores() - 2L)
#
#   old_plan <- future::plan()
#   on.exit(future::plan(old_plan), add = TRUE)
#   future::plan(future::multisession, workers = workers)
#
#   progressr::handlers(progressr::handler_rstudio())
#
#   progressr::with_progress({
#     p <- progressr::progressor(along = 1:n_sim)
#
#     mats <- furrr::future_map(
#       1:n_sim,
#       function(s) {
#         p(sprintf("sim %d / %d", s, n_sim))
#         one_sim(
#           s = s,
#           I = I, mu = mu, tau2 = tau2,
#           beta_m1_grid = beta_m1_grid,
#           model_list = model_list,
#           beta_m2 = beta_m2, beta_m3_levels = beta_m3_levels,
#           beta_c1 = beta_c1, beta_c2 = beta_c2, beta_c3 = beta_c3,
#           se_fun = se_fun
#         )
#       },
#       .options = furrr::furrr_options(seed = TRUE)
#     )
#
#     # stack into array [beta, model, sim]
#     simplify2array(mats)
#   })
# }
# progressr::handlers(progressr::handler_rstudio())
#
# # run it
# pvals_m1 <- run_power_parallel(n_sim)
#
# # power
# power_mat <- apply(pvals_m1, c(1, 2), \(v) mean(v < 0.05, na.rm = TRUE))
# power_mat
```
