---
title: "Gemini Classification & Inter-Rater Reliability"
subtitle: "Second LLM coder (Gemini 3.1 Pro) + Cohen's Kappa"
date: today
format:
  html:
    toc: true
    code-fold: true
execute:
  eval: true
---

## Setup

```{r}
#| label: gemini-setup
#| message: false

library(ellmer)
library(jsonlite)
library(here)
library(dplyr)
library(readr)
library(glue)
library(irr)

# Allow duplicate chunk labels when purling the pipeline
options(knitr.duplicate.label = "allow")

# Source pipeline functions for schema and prompts
pipeline_qmd <- here::here(
  "field_classification",
  "article_classification_pipeline - integrated.qmd"
)
pipeline_r <- knitr::purl(pipeline_qmd, output = tempfile(fileext = ".R"), quiet = TRUE)
source(pipeline_r, local = TRUE)
```

## Load Validation Articles

```{r}
#| label: load-articles

articles <- read_csv(
  here::here("field_classification", "validation_articles.csv"),
  show_col_types = FALSE
)

cat(sprintf("Articles to classify: %d\n", nrow(articles)))
```

## Classify with Gemini 3.1 Pro

Use the same cluster prompt as the main pipeline, but with Gemini instead of gpt-oss-120b.

```{r}
#| label: gemini-classify

create_gemini_cluster_chat <- function() {
  system_prompt <- glue(cluster_prompt_template,
                        cluster_schema = cluster_schema_json)
  chat_google_gemini(
    system_prompt = system_prompt,
    model = "gemini-3.1-pro-preview",
    credentials = function() Sys.getenv("GEMINI_API_KEY"),
    params = params(temperature = 0),
    echo = "none"
  )
}

n <- nrow(articles)
gemini_cluster_ids <- character(n)
gemini_confidences <- numeric(n)
gemini_secondary_ids <- character(n)
gemini_secondary_confs <- numeric(n)

cat(sprintf("Classifying %d articles with Gemini 3.1 Pro (sequential)...\n", n))

for (i in seq_len(n)) {
  if (i %% 10 == 0 || i == n) cat(sprintf("  %d / %d\n", i, n))

  article_json <- toJSON(list(
    article_title = articles$title[i] %||% "",
    article_abstract = articles$abstract[i] %||% "",
    journal_title = articles$source_display_name[i] %||% ""
  ), auto_unbox = TRUE) |> as.character() |> clean_json_text()

  tryCatch({
    chat <- create_gemini_cluster_chat()
    chat$chat(article_json)
    turns <- chat$get_turns()
    content <- turns[[2]]@text

    # Gemini may wrap JSON in markdown code fences
    content <- gsub("^```json\\s*", "", content)
    content <- gsub("^```\\s*", "", content)
    content <- gsub("\\s*```$", "", content)

    parsed <- tryCatch(fromJSON(content),
      error = function(e) list(cluster_id = NA_character_, confidence = NA_real_,
                               secondary_cluster_id = NA_character_,
                               secondary_confidence = NA_real_))

    gemini_cluster_ids[i] <- parsed$cluster_id %||% NA_character_
    gemini_confidences[i] <- parsed$confidence %||% NA_real_
    gemini_secondary_ids[i] <- parsed$secondary_cluster_id %||% NA_character_
    gemini_secondary_confs[i] <- parsed$secondary_confidence %||% NA_real_
  }, error = function(e) {
    message("  Error on article ", i, ": ", e$message)
    gemini_cluster_ids[i] <<- NA_character_
    gemini_confidences[i] <<- NA_real_
    gemini_secondary_ids[i] <<- NA_character_
    gemini_secondary_confs[i] <<- NA_real_
  })
}

gemini_results <- articles |>
  select(doi) |>
  mutate(
    gemini_cluster_id = gemini_cluster_ids,
    gemini_confidence = gemini_confidences,
    gemini_secondary_cluster_id = gemini_secondary_ids,
    gemini_secondary_confidence = gemini_secondary_confs
  )

write_csv(gemini_results, here::here("field_classification", "validation_gemini_classified.csv"))
cat("Saved Gemini classifications to validation_gemini_classified.csv\n")
```

## Gemini Classification Summary

```{r}
#| label: gemini-summary

cat("=== Gemini Subject Area Distribution ===\n")
gemini_results |>
  count(gemini_cluster_id, sort = TRUE) |>
  print(n = Inf)

cat(sprintf("\n%d articles with NA cluster_id\n",
            sum(is.na(gemini_results$gemini_cluster_id))))

cat("\n=== Gemini Confidence Distribution ===\n")
summary(gemini_results$gemini_confidence)
```

## Inter-Rater Reliability: gpt-oss-120b vs Gemini 3.1 Pro

```{r}
#| label: kappa-llms

# Load gpt-oss-120b results
gpt_results <- read_csv(
  here::here("field_classification", "validation_classified.csv"),
  show_col_types = FALSE
)

# Merge on DOI
comparison <- inner_join(
  gpt_results |> select(doi, gpt_cluster_id = cluster_id),
  gemini_results |> select(doi, gemini_cluster_id),
  by = "doi"
) |>
  filter(!is.na(gpt_cluster_id) & !is.na(gemini_cluster_id))

cat(sprintf("Articles with both classifications: %d\n\n", nrow(comparison)))

# Agreement rate
agreement <- mean(comparison$gpt_cluster_id == comparison$gemini_cluster_id)
cat(sprintf("Raw agreement: %.1f%% (%d / %d)\n",
            agreement * 100, sum(comparison$gpt_cluster_id == comparison$gemini_cluster_id),
            nrow(comparison)))

# Cohen's Kappa
kappa_result <- kappa2(comparison[, c("gpt_cluster_id", "gemini_cluster_id")])
cat(sprintf("\nCohen's Kappa: %.3f\n", kappa_result$value))
cat(sprintf("z = %.2f, p = %s\n", kappa_result$statistic,
            format.pval(kappa_result$p.value, digits = 3)))

cat("\n=== Disagreements ===\n")
disagreements <- comparison |> filter(gpt_cluster_id != gemini_cluster_id)
cat(sprintf("%d disagreements\n", nrow(disagreements)))
if (nrow(disagreements) > 0) {
  disagreements |>
    left_join(articles |> select(doi, title, source_display_name), by = "doi") |>
    select(doi, title, source_display_name, gpt_cluster_id, gemini_cluster_id) |>
    print(n = 30)
}
```

## Full Three-Way Agreement (after manual coding)

Load the manual coding results and compute Fleiss' Kappa across all three raters.

```{r}
#| label: kappa-three-way
#| eval: false

# Load manual coding (download from the HTML app)
manual <- read_csv(
  here::here("field_classification", "manual_classification.csv"),
  show_col_types = FALSE
) |>
  filter(cluster_id != "")

# Merge all three
three_way <- comparison |>
  inner_join(manual |> select(doi, manual_cluster_id = cluster_id), by = "doi") |>
  filter(!is.na(manual_cluster_id) & manual_cluster_id != "")

cat(sprintf("Articles with all three classifications: %d\n\n", nrow(three_way)))

# Pairwise kappas
cat("=== Pairwise Cohen's Kappa ===\n")
k_gpt_gemini <- kappa2(three_way[, c("gpt_cluster_id", "gemini_cluster_id")])
k_gpt_manual <- kappa2(three_way[, c("gpt_cluster_id", "manual_cluster_id")])
k_gemini_manual <- kappa2(three_way[, c("gemini_cluster_id", "manual_cluster_id")])

cat(sprintf("gpt-oss-120b vs Gemini:  %.3f\n", k_gpt_gemini$value))
cat(sprintf("gpt-oss-120b vs Manual:  %.3f\n", k_gpt_manual$value))
cat(sprintf("Gemini vs Manual:        %.3f\n", k_gemini_manual$value))

# Fleiss' Kappa (all three raters)
# irr::kappam.fleiss expects a matrix: rows = subjects, cols = raters
ratings_matrix <- three_way |>
  select(gpt_cluster_id, gemini_cluster_id, manual_cluster_id) |>
  as.matrix()

fleiss_result <- kappam.fleiss(ratings_matrix)
cat(sprintf("\nFleiss' Kappa (3 raters): %.3f\n", fleiss_result$value))
cat(sprintf("z = %.2f, p = %s\n", fleiss_result$statistic,
            format.pval(fleiss_result$p.value, digits = 3)))

# Three-way agreement
all_agree <- mean(three_way$gpt_cluster_id == three_way$gemini_cluster_id &
                  three_way$gpt_cluster_id == three_way$manual_cluster_id)
cat(sprintf("\nThree-way raw agreement: %.1f%%\n", all_agree * 100))
```
